# Batch Correction

As soon as a project contains multiple samples, donors, or experimental runs, batch effects become impossible to ignore. If they are not handled properly, cells separate by technical origin instead of biology, and every downstream result becomes harder to trust.

This chapter shows how `sclet` handles multi-sample workflows in a way that remains explicit and inspectable. We start from preprocessing multiple datasets, then move into integration and corrected views that can be reused consistently by downstream steps.

## Load datasets

We load the PBMC 3k and 4k datasets using `TENxPBMCData`. Note that these datasets use Ensembl IDs as row names by default. To facilitate downstream analysis (e.g., cell type annotation with `SingleR` which requires Gene Symbols matching the reference), we map the Ensembl IDs to Gene Symbols. We use `scuttle::uniquifyFeatureNames()` to perform this mapping while ensuring that the resulting row names remain unique.

```{r pbmc3k4k}
library(TENxPBMCData)
library(scuttle)

pbmc3k <- TENxPBMCData(dataset = "pbmc3k")
# Replace Ensembl IDs with Gene Symbols (ensuring uniqueness) for compatibility with downstream tools
rownames(pbmc3k) <- scuttle::uniquifyFeatureNames(rownames(pbmc3k), rowData(pbmc3k)$Symbol_TENx)

pbmc4k <- TENxPBMCData(dataset = "pbmc4k")
rownames(pbmc4k) <- scuttle::uniquifyFeatureNames(rownames(pbmc4k), rowData(pbmc4k)$Symbol_TENx)
```



## Preprocess
```{r pbmc3k4k-preprocess}
library(sclet)

sce_process <- function(sce) {
  sce <- QCMetrics(sce)
  sce[["percent.mt"]] <- PercentageFeatureSet(sce, "^MT-")
  keep <- colData(sce)$nFeature_RNA > 200 &
    colData(sce)$nFeature_RNA < 2500 &
    colData(sce)$percent.mt < 5
  sce <- sce[, keep, drop = FALSE]
  sce <- NormalizeData(sce)
  sce <- FindVariableFeatures(sce)
  return(sce)
}

pbmc3k <- sce_process(pbmc3k)
pbmc4k <- sce_process(pbmc4k)

pbmc_list <- list(pbmc3k = pbmc3k, pbmc4k = pbmc4k)
```

## Integration Workflow

`sclet` provides a unified `RunIntegration` entry point to perform multi-backend data integration (e.g., `fastMNN`, `Harmony`). It automatically manages the resulting representations (layers or reductions) and registers an active integration state so that downstream tools can consume them seamlessly.

```{r batch-remove}
# Merge datasets
combined_pbmc <- sce_merge(pbmc_list)

# Integration using fastMNN (default)
rm_batch_pbmc <- RunIntegration(combined_pbmc, method = "fastMNN", batch = "batch")

# RunIntegration registers a logical "corrected" layer for fastMNN
DefaultLayer(rm_batch_pbmc)
Layers(rm_batch_pbmc)

# and records a typed integration state
has_integration(rm_batch_pbmc)
get_integration(rm_batch_pbmc)
```

Alternatively, you can use `Harmony` or `scVI` (Deep Learning via `basilisk`), which operate on the latent space and return a corrected reduction rather than an expression layer:

The Harmony example is executed during book rendering, which means it also serves as a regression test in CI. The `scVI` call is also executed when the Python-backed `basilisk` environment is available, giving the book a minimal end-to-end test of the deep generative integration path.

```{r run-harmony-scvi}
# Reduction-based integration methods expect an existing latent space.
if (!"PCA" %in% SingleCellExperiment::reducedDimNames(combined_pbmc)) {
  combined_pbmc <- RunPCA(
    combined_pbmc,
    subset_row = VariableFeatures(combined_pbmc),
    layer = "logcounts"
  )
}

# Integration using Harmony
rm_harmony_pbmc <- RunIntegration(combined_pbmc, method = "Harmony", batch = "batch")

# Harmony creates a reduction named "HARMONY" and sets it as the active reduction
DefaultReduction(rm_harmony_pbmc)

```

```{r run-scvi, eval=eval_chunks && tolower(Sys.getenv("SCLET_BOOKDOWN_EVAL_PYTHON", "false")) %in% c("true", "1", "yes") && requireNamespace("basilisk", quietly = TRUE) && requireNamespace("reticulate", quietly = TRUE)}
# Integration using scVI (Requires Python basilisk environment)
rm_scvi_pbmc <- RunIntegration(combined_pbmc, method = "scVI", batch = "batch")
DefaultReduction(rm_scvi_pbmc) # Returns "scVI"
```

## Clustering

```{r clustering-batch}
sce_clustering <- function(sce, layer = DefaultLayer(sce), dims = 1:10) {
  # For uncorrected data, we run PCA first
  if (!"PCA" %in% SingleCellExperiment::reducedDimNames(sce)) {
      sce <- RunPCA(sce, subset_row = VariableFeatures(sce), layer = layer)
  }
  
  # Downstream tools automatically perceive the integration state
  # (e.g., FindNeighbors will use "HARMONY" if it is the active reduction)
  sce <- FindNeighbors(sce, dims = dims)
  sce <- FindClusters(sce)
  sce <- RunUMAP(sce, dims = dims, layer = layer)
  return(sce)
}

combined_pbmc <- sce_clustering(combined_pbmc, layer = "logcounts")

rm_batch_pbmc <- sce_clustering(rm_batch_pbmc, layer = "corrected")

```

The important point is that downstream analysis functions can now consume the integration outputs directly, without requiring users to manually point to the `reconstructed` assay or the `HARMONY` reduction. `RunIntegration()` sets up the active pointers in the `analysis-state` contract.

The merge step is also part of that story now. `sce_merge()` stores a non-active `merged_inputs` integration record, and `BatchRemover()` points back to it. This gives the merged object and the corrected object a connected provenance chain instead of two unrelated transformations.

You can inspect how this provenance is carried into downstream states:

```{r batch-provenance, eval = eval_chunks}
get_integration(rm_batch_pbmc)
get_graph(rm_batch_pbmc)$inputs$integration
get_analysis_context(rm_batch_pbmc)$active
```

## Visualization


```{r batch-umap, fig.width=13.5, fig.height=12}
library(ggplot2)
library(aplot)

p1 <- DimPlot(combined_pbmc, reduction="UMAP", mapping=aes(color=batch)) 
p2 <- DimPlot(rm_batch_pbmc, reduction="UMAP", mapping=aes(color=batch)) 
p3 <- plot_list(Before=p1, After=p2)

p4 <- DimPlot(rm_batch_pbmc, reduction="UMAP")

plot_list(p3, p4, ncol=1, heights=c(1, 2))
```

## Pseudobulk Differential Expression

sclet now supports pseudobulk analysis to perform robust differential expression between conditions/samples.

This pseudobulk example is also shown without execution, because real use requires meaningful biological replicates and an analysis-specific design formula. In other words, the missing piece here is experimental design rather than software support: a documentation build can fabricate toy labels, but it cannot invent biologically valid replicate structure.

```{r pseudobulk-de, eval = FALSE}
# 1. Aggregate counts by cluster and batch (or sample)
# This creates a pseudobulk SummarizedExperiment
pb <- AggregateExpression(rm_batch_pbmc, group_by = c("label", "batch"))

# 2. Perform DE analysis (using DESeq2)
# Here we test for differences between batches within a specific cluster (e.g., cluster 1)
# Note: Real analysis should have biological replicates, here we use batch as an example.

# Filter for cluster 1 only
pb_cluster1 <- pb[, pb$label == "1"]

# Run DE: test for batch effect
# design formula: ~ batch
# contrast: batch pbmc4k vs pbmc3k
res <- FindMarkers_pseudobulk(pb_cluster1, design = ~ batch, contrast = c("batch", "pbmc4k", "pbmc3k"))

head(res)
```





