# Cell-Cell Communication

Once cell identities have been assigned, one common next question is how these populations may interact with one another. Ligand-receptor analysis is a natural way to move from isolated cell types toward a more explicit view of the tissue microenvironment.

This chapter introduces the `sclet` unified interface for Cell-Cell Communication (CCI), with an emphasis on keeping communication analysis aligned with the same layer-aware and state-aware conventions used elsewhere in the package.

## Input

The recommended entry point is `RunCCI()`. It requires a `SingleCellExperiment` object that includes cell clustering information. By default, it delegates to `CellChat` as the computational backend.

The `group` parameter should reference a variable within the `SingleCellExperiment` object that holds the clustering data.

`RunCCI()` follows the same layer contract as the rest of `sclet`. In most RNA workflows, `layer = "logcounts"` is the safest explicit choice. If you omit it, the function resolves from `DefaultLayer(sce)` and avoids scaled layers when necessary.

Users can add their own clustering using the following commands:

```r
colData(sce)$group_name <- group
# OR
sce[["group_name"]] <- group
```

## Running CCI

Wrapper function for unified CCI analysis. Behind the scenes, it extracts standardized ligand-receptor interactions into the state tree. By default, `keep_object = FALSE` to prevent the `SingleCellExperiment` object from bloating with massive third-party objects. 

Here we set `keep_object = TRUE` purely because we want to demonstrate the legacy CellChat visualization in the final section of this chapter.

```{r}
library(sclet)

# use the bundled pancreas SCE demo object
sce <- readRDS("data/pancreas_sub_sce.rds")
sce <- RunCCI(sce, method = "CellChat", group = "CellType", layer = "logcounts", name = "cellchat_main", species = "mouse", keep_object = TRUE)
```

After running `RunCCI()`, the updated SCE keeps a structured communication record. This is useful when you want downstream code to check `has_cci(sce)` or inspect the standardized interactions data frame through `get_cci(sce)`.

```{r}
# Extract the standardized interactions data.frame
cci_df <- get_cci(sce)
head(cci_df)
```

If you keep multiple communication runs in the same object, give them different `name` values and inspect a specific record with `get_cci(sce, id = "your_id")`.

## Using Alternative Backends (CellPhoneDB & NicheNet)

`sclet` acts as a unified adapter. You can easily switch to other popular CCI backends without changing your downstream data extraction or visualization code.

### CellPhoneDB

CellPhoneDB is an established method that relies on a curated database of ligands and receptors. Through `sclet`, it runs directly via Python integration (`basilisk`):

```{r, eval=FALSE}
# Run CellPhoneDB analysis
sce <- RunCCI(
    sce, 
    method = "CellPhoneDB", 
    group = "CellType", 
    species = "mouse", # will attempt ortholog mapping internally
    name = "cellphonedb_main",
    threads = 2
)

# Extract standard results
cpdb_df <- get_cci(sce, id = "cellphonedb_main")
head(cpdb_df)
```

### NicheNet

While CellChat and CellPhoneDB focus on ligand-receptor pairs, NicheNet takes it a step further by predicting which ligands from sender cells regulate specific gene sets (e.g., DEGs) in receiver cells.

NicheNet requires pre-downloaded prior models (ligand-target matrix, etc.).

```{r, eval=FALSE}
# Load NicheNet prior models (you must download these first)
ligand_target_matrix <- readRDS("ligand_target_matrix.rds")
lr_network <- readRDS("lr_network.rds")
weighted_networks <- readRDS("weighted_networks.rds")

# Run NicheNet to find ligands from Endocrine cells affecting Epithelial cells
sce <- RunCCI(
    sce,
    method = "NicheNet",
    group = "CellType",
    receiver = "Epithelial",
    sender = "Endocrine",
    # If genes_oi is NULL, sclet automatically finds DEGs for the receiver cluster
    ligand_target_matrix = ligand_target_matrix,
    lr_network = lr_network,
    weighted_networks = weighted_networks,
    name = "nichenet_main",
    top_n_ligands = 20
)

nichenet_df <- get_cci(sce, id = "nichenet_main")
head(nichenet_df)
```

## Visualization

Instead of relying heavily on third-party specific visualization functions, `sclet` provides native `ggplot2` and `circlize` based visualization functions that directly consume the standardized CCI data layer. This decouples your analysis from specific backend plotting idiosyncrasies.

```{r cci-bubble, fig.width=6, fig.height=6}
# Native Bubble plot for specific cell pairs
sc_cci_bubble(sce, sources = c("Ductal", "Endocrine"), targets = c("Ductal", "Endocrine"))
```

```{r cci-chord, fig.width=6, fig.height=6}
# Native Chord diagram for specific pathways
sc_cci_chord(sce, signaling = c("PTN"))
```

### Legacy CellChat Visualization

If you still need the full power of CellChat's specific visualizations, you can retrieve the original `CellChat` object and use it directly.

```{r cci-legacy-setup}
# Retrieve the original CellChat object if you need legacy plots
cci_obj <- get_cellchat(sce, element = "object")
```

```{r cci-1, fig.width=15, fig.height=8}
library(aplot)
groupSize <- as.numeric(table(cci_obj@idents))
plot_list(
    ~CellChat::netVisual_circle(cci_obj@net$count, 
        vertex.weight = groupSize, weight.scale = T, 
        label.edge= F, title.name = "Number of interactions"),
    ~CellChat::netVisual_circle(cci_obj@net$weight, 
        vertex.weight = groupSize, weight.scale = T, 
        label.edge= F, title.name = "Interaction weights/strength")
)
```

```{r cci-2, fig.width=7.5, fig.height=8}
pathways.show <- c("PTN")
CellChat::netVisual_aggregate(cci_obj, signaling = pathways.show, layout = "circle")
```

```{r cci-3, fig.width=7.5, fig.height=8}
CellChat::netVisual_aggregate(cci_obj, signaling = pathways.show, layout = "chord")
```


```{r cci-4, fig.width=6, fig.height=6}
CellChat::netVisual_heatmap(cci_obj, signaling = pathways.show, color.heatmap = "Reds")
```

```{r cci-5, fig.width=6, fig.height=6}
group.cellType <- c(rep("Epithelial", 3), rep("Endocrine", 2)) # grouping cell clusters into Epithelial and Endocrine cells
names(group.cellType) <- levels(cci_obj@idents)
CellChat::netVisual_chord_cell(cci_obj, signaling = pathways.show, group = group.cellType, title.name = paste0(pathways.show, " signaling network"))
```


```{r cci-6, fig.width=6, fig.height=6}
CellChat::netAnalysis_contribution(cci_obj, signaling = pathways.show)
```


```{r cci-7, fig.width=6, fig.height=6}
pairLR.pathway <- CellChat::extractEnrichedLR(cci_obj, signaling = pathways.show, geneLR.return = FALSE)
LR.show <- pairLR.pathway[1,]
vertex.receiver = seq(1,4)
# CellChat::netVisual_individual(cci_obj, signaling = pathways.show,  pairLR.use = LR.show, vertex.receiver = vertex.receiver)
CellChat::netVisual_individual(cci_obj, signaling = pathways.show, pairLR.use = LR.show, layout = "circle")
```


```{r cci-8, fig.width=6, fig.height=6}
CellChat::netVisual_individual(cci_obj, signaling = pathways.show, pairLR.use = LR.show, layout = "chord")
```

```{r cci-9, fig.width=5, fig.height=6}
CellChat::netVisual_bubble(cci_obj, sources.use = 1:2, targets.use = c(3:5), remove.isolate = FALSE)
```


