# Cell Identity and Reference Mapping {#reference-mapping}

This mainline answers the most common post-clustering question: what are these cells, and how confidently can we connect them to a known reference atlas?

`sclet` keeps manual annotation, lightweight label transfer, and reference-driven mapping in the same narrative. The goal is to let users think in terms of one annotation workflow rather than a choice between disconnected backend names.

## Manual annotation

Please refer to the [Cell cluster annotation](#manual-annotation) session. 


## Automatic annotation
    
The low-level `SingleR::SingleR()` call is still useful for understanding the underlying method. In daily `sclet` workflows, however, the recommended entry point is `RunSingleR()` or the semantic wrapper `RunReferenceMapping(method = "SingleR")`, because these paths record the annotation result and its provenance back into the unified state layer.

At the moment, the practical backend stack is intentionally simple:

1. `SingleR` as the main reference annotation / mapping entry
2. `KNN` as a lightweight baseline mapping backend
3. `Symphony` as a future projection-oriented backend that is reserved but not yet implemented

In this chapter, both the low-level `SingleR()` workflow and the `sclet` analysis examples are executed during book rendering when the required Bioconductor dependencies are available. This keeps the chapter useful as documentation while also exercising the annotation layer in CI.

## Reference mapping mainline workflow

If your real question is "what does this query dataset correspond to in a reference atlas?", `RunReferenceWorkflow()` is now the semantic entry point for that mainline. It wraps the reference-mapping step as one reusable analysis unit instead of forcing you to think first about whether the backend is `SingleR` or `KNN`.

### Lightweight Reference Mapping (KNN)

If you have a reference `SingleCellExperiment` with cell type annotations, you can use it to map labels to your query dataset quickly:

```{r run-knn-predict, eval = eval_chunks && requireNamespace("BiocNeighbors", quietly = TRUE)}
set.seed(1)
ref_idx <- sample(seq_len(ncol(pbmc)), 500)
ref_sce <- pbmc[, ref_idx]
query_sce <- pbmc[, -ref_idx]

ref_sce$label <- as.character(SingleCellExperiment::colLabels(ref_sce))

query_sce <- RunReferenceMapping(
    object = query_sce,
    ref = ref_sce,
    labels = "label",
    method = "KNN",
    layer = "logcounts",
    k = 5,
    name = "knn_demo"
)

# Inspect the recorded reference-mapping state
get_mapping(query_sce)

# Visualize the query and reference in the same projection space
ProjectionPlot(query_sce, ref_sce)
```

### Run SingleR with sclet

```{r run-singleR-sclet, eval = eval_chunks && requireNamespace("SingleR", quietly = TRUE) && requireNamespace("celldex", quietly = TRUE)}
library(sclet)
hpca <- celldex::HumanPrimaryCellAtlasData()

pbmc2 <- RunReferenceMapping(
    pbmc,
    ref = hpca,
    labels = hpca$label.main,
    method = "SingleR",
    layer = "logcounts",
    name = "hpca_main"
)

# inspect the recorded annotation state
get_annotation(pbmc2, id = "hpca_main")

# inspect the recorded reference-mapping state
get_mapping(pbmc2, id = "hpca_main")

# check the stored label columns
head(colnames(SummarizedExperiment::colData(pbmc2)))

# summarize transferred labels across query groups
plot_reference_label_transfer_heatmap(
    pbmc2,
    group.by = "colLabels",
    id = "hpca_main"
)

# inspect confidence / score distributions across transferred labels
plot_reference_label_confidence(
    pbmc2,
    id = "hpca_main"
)
```

```{r run-reference-workflow, eval = eval_chunks && requireNamespace("SingleR", quietly = TRUE) && requireNamespace("celldex", quietly = TRUE)}
pbmc_ref <- RunReferenceWorkflow(
    pbmc,
    ref = hpca,
    labels = hpca$label.main,
    method = "SingleR",
    layer = "logcounts",
    name = "reference_main"
)

plot_reference_label_transfer_heatmap(
    pbmc_ref,
    group.by = "colLabels",
    id = "reference_main_mapping"
)

plot_reference_label_confidence(
    pbmc_ref,
    id = "reference_main_mapping"
)
```


### Run SingleR

```{r run-singleR, eval = eval_chunks && requireNamespace("SingleR", quietly = TRUE) && requireNamespace("celldex", quietly = TRUE)}

hpca <- celldex::HumanPrimaryCellAtlasData()
hpca
res <- SingleR::SingleR(
    test = SummarizedExperiment::assay(pbmc, "logcounts"),
    ref = hpca,
    labels = hpca$label.main
)
head(res)


dice <- celldex::DatabaseImmuneCellExpressionData()
res2 <- SingleR::SingleR(
    test = SummarizedExperiment::assay(pbmc, "logcounts"),
    ref = dice,
    labels = dice$label.main
)
head(res2)
```

When you use the low-level `SingleR()` call directly, you are responsible for deciding which matrix to pass. In contrast, `RunReferenceMapping(method = "SingleR")` follows the current `sclet` layer contract and resolves that choice from `DefaultLayer(pbmc)` unless you override it. If you provide `name = "hpca_main"`, both the annotation record and the mapping record are stored under that id, and the predicted labels are written to run-specific columns such as `hpca_main_labels`.

If the current layer comes from a corrected integration workflow, both the annotation state and the mapping state record that provenance. In practice, this means `get_annotation(pbmc)$inputs$integration` and `get_mapping(pbmc)$inputs$integration` can tell you whether the label transfer was performed on top of a corrected representation rather than the raw normalized layer.

### Annotate with SingleR result

```{r annotation, eval = eval_chunks && exists("res") && exists("res2")}
pbmc[['hpca_label']] <- res$labels

pbmc[['dice_label']] <- res2$labels
```


### Comparision with manual annotation result

```{r annotation-comparison, fig.width=16, fig.height=6, eval = eval_chunks && exists("res") && exists("res2")}
# Ensure manual labels are available in colData for comparison
pbmc$label <- SingleCellExperiment::colLabels(pbmc)

x <- colData(pbmc)[, c("label", "hpca_label", "dice_label")]

plot_list(
    manual = sc_dim(pbmc, reduction="UMAP"),    
    hpca = sc_dim(pbmc, reduction="UMAP", mapping=aes(color=hpca_label)),
    dice = sc_dim(pbmc, reduction="UMAP", mapping=aes(color=dice_label)),
    ncol = 3) & 
sc_dim_geom_label(geom = ggrepel::geom_text_repel) & 
theme(legend.position='inside')

table(x[,c(1,2)])

table(x[,c(1,3)])
```

