# Cell Fate and Dynamic Processes {#cell-fate}

This mainline shifts the analysis story from static identity to directional biology. Once cells have been placed in a baseline manifold, the next question is often whether they lie along a continuum, where terminal states sit, and which transitions are supported by the data.

`sclet` gathers diffusion-based embeddings, lineage inference, velocity, and fate-oriented methods under the same workflow umbrella. The point is not just to expose more algorithms, but to make the dynamics route readable and reusable as one coherent chapter of analysis.

```{r fate-setup, include=FALSE}
# Velocity and CellRank are Python-backed through basilisk/velociraptor.
# The main bookdown build reads cached RDS objects from data/ instead of
# provisioning Python environments during render.
velocity_cache <- "data/hermann-sperm-velo.rds"
dynamics_cache <- "data/hermann-sperm-dynamics.rds"
velocity_cache_ready <- eval_chunks && file.exists(velocity_cache)
#dynamics_cache_ready <- eval_chunks && file.exists(dynamics_cache)

dynamics_cache_ready <- TRUE
```

## Trajectory mainline workflow

If you do not want to stitch together `RunDiffusionMap()`, `RunSlingshot()`, `RunVelocity()`, and `RunCellFate()` by hand every time, `sclet` now provides `RunTrajectoryWorkflow()` as the semantic entry point for this entire analysis mainline.

The point of this wrapper is not to hide the underlying methods. The point is to make the recommended analysis path easier to follow, easier to document, and easier to rerun later.

```{r trajectory-mainline, eval = eval_chunks}
panc_main <- readRDS("data/pancreas_sub_sce.rds")
panc_main <- RunTrajectoryWorkflow(
  panc_main,
  group = "SubCellType",
  reduction = "UMAP",
  run_diffusion_map = TRUE,
  run_slingshot = TRUE,
  run_velocity = "never",
  run_fate = "never",
  name = "pancreas_main"
)
```

```{r trajectory-mainline-overview, fig.width=11, fig.height=8, eval = eval_chunks}
plot_trajectory_overview(
  panc_main,
  id = "pancreas_main",
  panels = c("lineage")
)
```




```{r}
panc <- readRDS("data/pancreas_sub_sce.rds")

panc <- RunSlingshot(sce = panc, group = "SubCellType", reduction = "UMAP")
```

`RunSlingshot()` writes its result back into the unified state layer, so downstream helpers such as `plot_lineage()` and `plot_pseudotime()` can consume the recorded trajectory object directly. If needed, you can also check `has_trajectory(panc)` or inspect `get_trajectory(panc)`.

If you keep multiple trajectory runs in the same object, give them different `name` values and inspect them explicitly with `get_trajectory(panc, id = "your_id")`.




### Lineage plot
```{r lineage_plot, fig.width=7, fig.height=8}
p1 <- plot_lineage(panc, group = "SubCellType") + 
  sc_dim_geom_label(
    geom = shadowtext::geom_shadowtext,
    color='black', 
    bg.color='white'
  )


## displayed selected lineages
p2 <- plot_lineage(panc, group = "SubCellType", lineages = c("Lineage1", "Lineage2"))  + 
  sc_dim_geom_label(
    geom = shadowtext::geom_shadowtext,
    color='black', 
    bg.color='white'
  )

plot_list(`All Lineages` = p1, `Selected Lineages` = p2, ncol=1)
```

### Pseudotime plot

```{r pseudotimeplot, fig.width=12, fig.height=10}
pp1 <- plot_pseudotime(panc, reduction = "UMAP", lineage = "Lineage1")
pp2 <- plot_pseudotime(panc, reduction = "UMAP", lineage = "Lineage2")
pp_all <- plot_pseudotime(panc, reduction = "UMAP")

plot_list(pp1, pp2, pp_all, design="AABB\nCCCC")
```


### Expression trends in different cell trajectories

```{r genecurve, fig.width=14, fig.height=8}
curve_genes <- c("Mrpl15", "Tram1", "Ncoa2")

## 默认选取smooth方法直接拟合得到的基因表达值结果来画图
p_curve <- plot_genecurve(panc, features = curve_genes, method = "smooth")

## 也可以用GAM计算预测的基因表达值来画图
p_curve1 <- plot_genecurve(panc, features = curve_genes, method = "gam")

plot_list(p_curve, p_curve1, ncol=1)
```



### Heatmap

```{r pseudoheatmap, fig.width=10, fig.height=16}
heatmap_genes <- head(unique(rownames(panc)), 10)

## 默认按照拟时间排序
hmp1 <- plot_pseudo_heatmap(panc, features = heatmap_genes, lineage = "Lineage2")

## 可以手动对基因或细胞设置聚类(cluster = TRUE)，此时会打乱pseudotime的排序
hmp2 <- plot_pseudo_heatmap(panc, features = heatmap_genes, lineage = "Lineage2", cluster_columns = TRUE)

## 对基因聚类
hmp3 <- plot_pseudo_heatmap(panc, features = heatmap_genes, lineage = "Lineage2", cluster_rows = TRUE, sort = FALSE)

## 默认排序
hmp4 <- plot_pseudo_heatmap(panc, features = heatmap_genes, lineage = "Lineage2", cluster_rows = FALSE, sort = TRUE)

## 组合起来看就能看到两个参数的设定对行顺序的影响
plot_list(hmp1, hmp2, hmp3, hmp4, ncol=1)
```



## Diffusion Map for complex trajectories {#diffusion-map}

For very continuous developmental systems, PCA and UMAP are often convenient but not always topology-faithful. `RunDiffusionMap()` adds a diffusion-based representation that can be tracked like any other reduction in `sclet`, which is useful when you want a more trajectory-aware geometry before moving on to downstream lineage tools.

```{r diffusion-map, eval = eval_chunks && requireNamespace("destiny", quietly = TRUE)}
panc_dm <- panc

if (!"logcounts" %in% SummarizedExperiment::assayNames(panc_dm) &&
    "counts" %in% SummarizedExperiment::assayNames(panc_dm)) {
  panc_dm <- NormalizeData(panc_dm)
}

if (!"PCA" %in% SingleCellExperiment::reducedDimNames(panc_dm)) {
  dm_features <- rownames(panc_dm)[seq_len(min(500, nrow(panc_dm)))]
  dm_layer <- if ("logcounts" %in% SummarizedExperiment::assayNames(panc_dm)) "logcounts" else "counts"
  panc_dm <- RunPCA(panc_dm, subset_row = dm_features, layer = dm_layer)
}

panc_dm <- RunDiffusionMap(panc_dm, n_components = 10, reduction = "PCA")
head(SingleCellExperiment::reducedDim(panc_dm, "DM")[, 1:3])
```

```{r diffusion-map-plot, fig.width=8, fig.height=6, eval = eval_chunks && requireNamespace("destiny", quietly = TRUE)}
DimPlot(panc_dm, reduction = "DM", mapping = ggplot2::aes(color = SubCellType))
```






The velocity and CellRank examples in this section use cached objects generated by the `update-data-velocity-fate` GitHub Actions workflow. This keeps the book render lightweight: Python-backed `basilisk`, `velociraptor`, and CellRank work is done in the data-update workflow, while the book only reads the resulting RDS files.

RNA metabolic process:

+ RNA is synthesized at a rate of alpha 
+ RNA is spliced to remove introns and form mature mRNA at rate of beta
+ Mature mRNA is degraded after functioning at rate of gamma

The disruption of the equilibrium between unspliced and spliced RNA indicates whether a gene is in an induced or repressed state. 

With RNA-seq, especially the Poly-A enrichment method, the propotion of unspliced RNA obtained is relatively small. Currently, there are also some techinques capable of enriching nascent RNA in single cells, such as [scSLAM-seq](https://www.nature.com/articles/s41586-019-1369-y) and s[cNT-seq](https://www.nature.com/articles/s41592-020-0935-4).


Although some researchers are skeptical about RNA velocity, it has received significant attention because it can automatically detect the direction of trajectories.


[scVelo](https://scvelo.readthedocs.io/) supports a full dynamical model and various of utility functions [@bergen_generalizing_2020]. It only supports Python and can be run in R using `r Biocpkg("velociraptor")`. 

### Example data

> This example is adopted from <https://www.bioconductor.org/packages/devel/bioc/vignettes/velociraptor/inst/doc/velociraptor.html>.


```{r velocity-data, eval = TRUE}
sce <- readRDS("data/hermann-sperm-dynamics.rds")
sce
```

```{r velocity-hvgs, eval = TRUE}
top_genes <- S4Vectors::metadata(sce)$sclet_bookdown_cache$top_genes
head(top_genes)
```

### Run velocity analysis

`sclet` wraps the RNA velocity inference via `RunVelocity()`. Under the hood, it delegates the heavy lifting to `scVelo` (managed safely via the `basilisk`/`velociraptor` Python sandbox) and smoothly registers the outputs to the `velocity` analysis state.

```{r velocity, eval = TRUE}
get_velocity(sce)
```

### Visualize velocity-based trajectory

Once the `velocity` state is recorded, visualizing it is a breeze. You can use `VelocityPlot()` to draw the dynamic stream or arrows over any active dimensionality reduction space (like UMAP or t-SNE).

```{r velocity-plot, eval = TRUE && requireNamespace("velociraptor", quietly = TRUE)}
# Project the cached velocity stream onto the UMAP embedding
VelocityPlot(sce, reduction = "UMAP")
```

## CellRank: Advanced Fate Mapping {#cellrank}

When standard trajectory tools (like Slingshot) struggle with complex, cyclic, or highly heterogeneous developmental processes, **CellRank** shines. CellRank elegantly combines directed RNA velocity data with undirected transcriptomic similarity to robustly infer terminal states and cellular fate probabilities.

`sclet` brings the full power of `CellRank` directly to your SCE object, again utilizing the isolated `basilisk` Python environment to prevent dependency conflicts.

### Run CellRank

`RunCellFate()` is the user-facing entry point for fate mapping in `sclet`. At the moment it dispatches to `CellRank`, automatically senses the upstream velocity run, and records the resulting terminal states, named fate probabilities, and lineage drivers in the shared state contract.

```{r cellrank-run, eval = TRUE}
# Inspect the cached CellRank analysis state
get_cellrank(sce, id = "cellrank")
```

The resulting terminal state assignments and directional fate probabilities are written directly into `colData(sce)`. In addition, `sclet` now provides fate-specific plotting helpers following the `plot_<module>_<type>` convention, so you do not need to manually reconstruct the plotting data frame after every run.

```{r cellrank-plot, fig.width=12, fig.height=5.5, eval = dynamics_cache_ready}
p_prob <- plot_fate_probability(sce, fate = 1, reduction = "UMAP", id = "cellrank")
p_term <- plot_fate_terminal_states(sce, reduction = "UMAP", id = "cellrank")

plot_list(`Fate Probability` = p_prob, `Terminal States` = p_term)
```

```{r cellrank-driver-trends, fig.width=11, fig.height=7, eval = dynamics_cache_ready}
cr <- get_cellrank(sce, id = "cellrank")
driver_lineage <- if (!is.null(cr$artifacts$fate_probability_names)) {
  cr$artifacts$fate_probability_names[[1]]
} else {
  cr$fate_probability_names[[1]]
}

plot_fate_driver_trends(sce, lineage = driver_lineage, top_n = 4)
```

## RegVelo with CellRank diagnostics {#regvelo-cellrank}

RegVelo extends RNA velocity inference by using a prior gene regulatory network, and in practice it is most useful when run on a GPU-enabled Python environment. The book build does not provision GPU resources, so this section follows the same cache-first pattern as the heavier CellRank examples above: the GPU step is shown as runnable user code, but the rendered book reads a precomputed `SingleCellExperiment` from `bookdown/data`.

The upstream run uses spliced and unspliced assays plus a prior GRN, writes the inferred velocity matrix back into the object as a named assay, and then hands that stored velocity result to CellRank. This code is intentionally not executed during book rendering.

```{r regvelo-gpu-run, eval=FALSE}
# Run this on a GPU server, not during book rendering.
sce <- RunRegVelo(
  sce,
  grn = prior_grn,
  reduction = "PCA",
  name = "regvelo_smoke",
  backend = "reticulate",
  python = "/path/to/gpu/venv/bin/python",
  max_epochs = 20,
  batch_size = 128
)

sce <- RunCellRank(
  sce,
  reduction = "PCA",
  cluster_key = "celltype_or_cluster",
  velocity_id = "regvelo_smoke",
  backend = "reticulate",
  python = "/path/to/gpu/venv/bin/python",
  name = "cellrank_regvelo_smoke"
)

saveRDS(sce, "bookdown/data/regvelo-smoke-cellrank.rds")
```

The cached object below was generated outside the book render. It lets the documentation demonstrate downstream inspection and plotting without rerunning PyTorch, RegVelo, or CellRank on the GitHub Actions worker.

```{r regvelo-cache, eval=TRUE}
regvelo_sce <- readRDS("data/regvelo-smoke-cellrank.rds")
get_velocity(regvelo_sce, id = "regvelo_smoke")$artifacts
```

```{r regvelo-cellrank-summary, eval=TRUE}
regvelo_diag <- CellRankSummary(regvelo_sce, id = "cellrank_regvelo_smoke")
regvelo_diag$summary
regvelo_diag$terminal_states
regvelo_diag$fate_probabilities
```

```{r regvelo-velocity-plots, fig.width=12, fig.height=5.5, eval=TRUE}
p_mag <- plot_velocity_magnitude(regvelo_sce, id = "regvelo_smoke", reduction = "PCA")
p_genes <- plot_top_velocity_genes(regvelo_sce, id = "regvelo_smoke", n = 15)

plot_list(`Velocity Magnitude` = p_mag, `Top Velocity Genes` = p_genes)
```

```{r regvelo-cellrank-plots, fig.width=12, fig.height=5.5, eval=TRUE}
p_regvelo_term <- plot_fate_terminal_states(
  regvelo_sce,
  id = "cellrank_regvelo_smoke",
  reduction = "PCA"
)
p_regvelo_fate <- plot_fate_probability(
  regvelo_sce,
  id = "cellrank_regvelo_smoke",
  fate = 1,
  reduction = "PCA"
)

plot_list(`Terminal States` = p_regvelo_term, `Fate Probability` = p_regvelo_fate)
```

```{r regvelo-fate-correlation, fig.width=7, fig.height=5.5, eval=TRUE}
VelocityFateCorrelation(
  regvelo_sce,
  velocity_id = "regvelo_smoke",
  trajectory_id = "cellrank_regvelo_smoke"
)

plot_velocity_fate_correlation(
  regvelo_sce,
  velocity_id = "regvelo_smoke",
  trajectory_id = "cellrank_regvelo_smoke"
)
```

In this small smoke-test object, CellRank detects only one terminal state and the single fate probability is effectively constant. That is still useful as a workflow check: it confirms that the RegVelo velocity assay can be consumed by the CellRank wrapper and downstream visualization. For biological interpretation, a real dataset should usually show multiple terminal states or nontrivial fate probability gradients.

## Workflow shell for velocity and fate

When spliced/unspliced assays are available, the same trajectory mainline can extend from lineage inference to velocity and fate mapping:

```{r trajectory-mainline-dynamic, eval=FALSE}
sce <- RunTrajectoryWorkflow(
  sce,
  group = "cluster",
  reduction = "PCA",
  run_diffusion_map = FALSE,
  run_slingshot = FALSE,
  run_velocity = "auto",
  run_fate = "auto",
  name = "germline_main"
)

plot_trajectory_overview(
  sce,
  id = "germline_main",
  panels = c("velocity", "terminal_states", "fate_probability")
)
```

## Velocity Latent Time

When scVelo is run in dynamical mode, it infers a latent time for each cell — a pseudotime ordering derived from the RNA velocity field itself, not from a separate trajectory method. `plot_velocity_latent_time()` colors cells on an embedding by this inferred temporal progression.

```{r velocity-latent, eval=FALSE}    
# requires scVelo dynamical mode latent time
plot_velocity_latent_time(
    sce,
    reduction = "UMAP",
    id = "germline_main_velocity"
)
```
