# Differential Expression & Pseudobulk {#differential-expression}

Finding marker genes is the bread and butter of single-cell analysis. But single-cell Differential Expression (DE) is notoriously tricky. Due to the vast number of cells (which inflates statistical power) and the zero-inflated nature of the data, standard single-cell DE often yields an overwhelming number of "significant" genes that turn out to be false positives or artifacts of biological replicates.

Reviewers at high-impact journals (like *Nature Methods*) are increasingly demanding **Pseudobulk** analysis to control for false discovery rates. `sclet` makes both standard DE and Pseudobulk DE incredibly straightforward.

In this chapter, the standard single-cell DE examples are executed during book rendering so they also serve as lightweight regression checks. The pseudobulk section now includes a small CI-friendly smoke test built from synthetic donor groupings, while the surrounding discussion still reflects the fact that real pseudobulk designs should be driven by actual biological replicates.

## Standard Single-Cell DE

You can find markers for all clusters using `RunDEtest()`, which wraps `FindMarkers` or `FindAllMarkers`. Importantly, `RunDEtest()` saves the result as an `analysis-state` inside the `SingleCellExperiment` object.

```{r detest, fig.width=10, fig.height=8, warning=FALSE, message=FALSE}
library(sclet)
# Assuming 'sce' has been processed and clustered
sce <- qs::qread("data/pbmc-3k-sce.qs")

# Run differential expression test and record the state
sce <- RunDEtest(sce, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

# Visualize the DE results using a Volcano plot
DEtestPlot(sce)
```

### The "What does this gene even do?" Problem

When you get a list of marker genes, you're often left staring at cryptic acronyms like `ZBTB16` or `KLRB1`. What do they actually do? 

`sclet` provides a handy utility called `gene_summary_table()` that automatically fetches the full name and biological summary of the genes from the `orgDb` databases.

```{r gene_summary, eval=requireNamespace("org.Hs.eg.db", quietly = TRUE)}
# Get the raw DE results from the state
de_results <- get_detest(sce)$results

# Take the top 5 markers
top_genes <- head(de_results$gene, 5)

# Fetch their biological summaries!
summary_df <- gene_summary_table(top_genes, orgDb = "org.Hs.eg.db")
print(summary_df)
```

This simple helper can save you hours of manually Googling gene names.

## The Gold Standard: Pseudobulk Analysis

To avoid the inflated p-values of single-cell DE, the current best practice is to aggregate the counts of all cells within the same biological replicate and cell type, creating a "pseudobulk" sample. You then run robust bulk RNA-seq tools (like `DESeq2` or `edgeR`) on these aggregated samples.

`sclet` handles this elegantly with two functions: `AggregateExpression()` and `FindMarkers_pseudobulk()`.

### Step 1: Aggregate Expression

In a real analysis, your `sce` object would usually already contain replicate metadata such as `donor_id` and `cell_type`. For a runnable book example, we create a simple donor column on the fly and reuse the current cluster labels as grouping labels.

```{r pseudobulk_agg, eval=eval_chunks}
# For a runnable example, we create a simple donor column.
cluster_id <- as.character(SingleCellExperiment::colLabels(sce))
if (length(cluster_id) == 0) {
    cluster_id <- as.character(Idents(sce))
}
sce$cluster_id <- cluster_id
sce$donor_id <- rep(paste0("donor", 1:4), length.out = ncol(sce))

# Aggregate counts by cluster and donor
pb_sce <- AggregateExpression(
    sce, 
    group_by = c("cluster_id", "donor_id"), 
    assay.type = "counts"
)

# The result is a much smaller SummarizedExperiment
# where each column is a "donor-cluster" pseudobulk sample.
dim(pb_sce)
```

### Step 2: Pseudobulk DE with DESeq2

Now, you can run `DESeq2` directly on this aggregated object to find markers robustly.

```{r pseudobulk_de, eval=eval_chunks && requireNamespace("DESeq2", quietly = TRUE)}
cluster_tab <- table(pb_sce$cluster_id)
target_clusters <- names(cluster_tab[cluster_tab >= 2])[1:2]
stopifnot(length(target_clusters) == 2)

pb_subset <- pb_sce[, pb_sce$cluster_id %in% target_clusters]
res <- FindMarkers_pseudobulk(
    pb_subset,
    design = ~ cluster_id,
    contrast = c("cluster_id", target_clusters[1], target_clusters[2])
)

head(res[, c("log2FoldChange", "pvalue", "padj")])
```

By leveraging `AggregateExpression()` and `FindMarkers_pseudobulk()`, you completely bypass the inflated p-value problem and ensure your biological findings are robust across actual replicates, not just individual cells.
