# Differential Expression & Enrichment {#enrichment}

`sclet` provides `RunDEtest` and `RunEnrichment` to seamlessly perform differential expression and functional enrichment analysis (GO/KEGG). These functions register their results into the `analysis-state` contract, making it easy for downstream visualization tools to consume them.

## Differential Expression

We can find markers for all clusters using `RunDEtest()`, which wraps `FindMarkers` or `FindAllMarkers` and saves the result as a state.

```{r detest, fig.width=10, fig.height=8}
library(sclet)
sce <- readRDS("data/pancreas_sub_sce.rds")

# Run differential expression test and record the state
sce <- RunDEtest(sce, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

# Inspect the recorded DE state
has_detest(sce)
head(get_detest(sce)$results)

# Visualize the DE results using a Volcano plot
DEtestPlot(sce)
```

## GO Enrichment

`RunEnrichment()` automatically reads the significant up-regulated markers from the active `detest` state.

```{r go_enrich, fig.width=10, fig.height=8}
# Ensure you have org.Mm.eg.db installed for Mouse data
# or org.Hs.eg.db for Human
library(org.Mm.eg.db) 

# Run enrichment analysis based on the DEtest state
sce <- RunEnrichment(
    sce, 
    db = "GO", 
    orgDb = "org.Mm.eg.db", 
    keyType = "SYMBOL", 
    ont = "BP"
)

# Visualize the enrichment results using EnrichmentPlot (wraps enrichplot::dotplot)
EnrichmentPlot(sce, showCategory=5)
```

## KEGG Enrichment

For KEGG, it is recommended to convert gene symbols to Entrez IDs first, or provide a column with Entrez IDs if available.

```{r kegg_enrich, eval=FALSE}
# Example if markers had ENTREZID
# sce <- RunEnrichment(sce, db = "KEGG", organism = "mmu")
# EnrichmentPlot(sce)
```
