# Gene Set Scoring {#geneset-scoring}

Often, relying on the expression of a single gene (like a traditional marker) introduces too much noise and dropout artifacts. If you want to know whether a cell population is undergoing "Hypoxia" or has an active "Cell Cycle", scoring the entire pathway or gene set is the most robust approach.

However, the algorithm landscape is fragmented: `AUCell`, `UCell`, `GSVA`... Each backend has a different output format. If you want to plot the results downstream, standardizing these formats can be a nightmare.

This is exactly the kind of problem `sclet` is designed to simplify. We provide the `RunGeneSetScoring()` interface, which unifies these backends and injects the scores directly into the state-aware workflow and `colData`, so the results remain easy to inspect and easy to reuse later.

## Multi-Backend Scoring in One Click

Let's define a few gene sets of interest and run the scoring on our PBMC dataset.

```{r geneset-scoring-setup}
library(sclet)
pbmc <- readRDS("data/pbmc-3k-sce.rds")

# Define signatures
my_signatures <- list(
  T_cell = c("CD3D", "CD3E", "CD8A", "CD4"),
  B_cell = c("CD79A", "MS4A1"),
  Myeloid = c("LYZ", "CD14", "FCGR3A")
)
```

You can decisively choose the backend that best fits your needs. `UCell` is rank-based, extremely fast, and highly suitable for single-cell data.

```{r geneset-scoring-run}
# Use UCell backend
pbmc <- RunGeneSetScoring(pbmc, gene_sets = my_signatures, method = "UCell", name = "UCell_Score")
```

## Reading and Visualizing Results

Where do the scores go after calculation?
1. By default, they are added as columns in the cellular metadata (`colData`).
2. The action is recorded in the `geneset_scoring` state machine.

Because the scores are now standard metadata, we can seamlessly reuse the `sclet` visualization ecosystem (like `FeatureDimPlot`):

```{r geneset-plot, fig.width=12, fig.height=5}
# Visualize the T_cell score distribution on the UMAP space
FeatureDimPlot(pbmc, features = "UCell_Score_T_cell")

# Inspect the scoring state provenance
get_geneset_scoring(pbmc)
```

## Advanced Pathway Discovery (altExp and RunKEGG)

When dealing with hundreds of pathways (like the entire KEGG database), dumping all scores into `colData` is messy. Instead, `sclet` allows storing these massive score matrices as an alternative experiment (`altExp`). This keeps your metadata clean while retaining perfect alignment with your cells.

Furthermore, `sclet` provides `RunKEGG()` as a one-stop wrapper. It automatically downloads pathways via `clusterProfiler`, scores the cells using `UCell`, stores the matrix in an `altExp`, and even identifies cell-type specific marker pathways using robust Wilcoxon rank-sum tests via `FindMarkerPathways()`.

```{r kegg-scoring, eval=FALSE}
# Run the complete KEGG workflow
# Note: Requires org.Hs.eg.db for human data
res <- RunKEGG(pbmc, species = "hsa", method = "UCell", group.by = "seurat_annotations")

# The updated SCE object with the KEGG altExp
pbmc <- res$sce

# The marker pathways data.frame
kegg_markers <- res$markers
head(kegg_markers)
```

You can also manually run differential testing on any pathway scores stored in an `altExp`:

```{r marker-pathways, eval=FALSE}
# If you ran RunGeneSetScoring with as_altExp = TRUE and name = "MyScores"
markers <- FindMarkerPathways(pbmc, altExp_name = "MyScores", group.by = "seurat_annotations")
```

**Bottom Line:**
Regardless of whether the underlying engine is `UCell` or `AUCell`, you can immediately plot the results with `FeatureDimPlot` or seamlessly run differential analysis. Life is too short to wrangle metadata formats.
