# Gene Regulatory Networks (pySCENIC) {#grn}

How do Transcription Factors (TFs) regulate target genes? Traditional differential expression analysis only tells you "which genes changed," but SCENIC tells you "who is issuing the commands."

If you have ever run the legacy R version of SCENIC, you undoubtedly remember its excruciatingly slow speed and the requirement to download tens of gigabytes of local databases. Today, the Python-based `pySCENIC` (powered by GRNBoost2) completely crushes the original R version in both speed and performance.

Rather than forcing users to leave the `SingleCellExperiment` workflow and manually manage a separate Python environment, `sclet` brings this functionality back into the same state-aware framework. Leveraging the `basilisk` isolated environment, it provides a native `sclet` entry point to `pySCENIC`: `RunSCENIC()`.

The examples in this chapter are shown but not executed during book rendering. A realistic `pySCENIC` workflow depends on external transcription-factor lists and cisTarget ranking databases, so it is more honest to present the code as a template that readers can adapt to their own resources. In practice, the limiting factor is not the Python environment itself, but the requirement to provide large database files that are version-matched to the species and genome build of your dataset.

## Environment and Data Preparation

You no longer need to manually configure complex Python environments or wrestle with `conda` conflicts. `sclet` securely prepares the sandbox in the background. You only need to provide three things:
1. Your single-cell expression matrix (SCE object)
2. A list of transcription factors (`.txt`)
3. The Motif database file (in `Feather` format)

For most human or mouse analyses, this also means checking that your TF list, motif annotation table, and cisTarget ranking database all come from the same release family before you start the run.

```{r grn-setup, eval=FALSE}
library(sclet)
# Assuming 'sce' is your pre-processed object

# Specify the paths to your downloaded databases
tfs_path <- "path/to/hs_hgnc_tfs.txt"
database_paths <- c(
  "path/to/hg38__refseq-r80__10kb_up_and_down_tss.mc9nr.feather"
)
motif_annotations_path <- "path/to/motifs-v9-nr.hgnc-m0.001-o0.0.tbl"
```

## One-Click Inference and Scoring

Next, call `RunSCENIC()`. All heavy lifting—GRNBoost2 inference, Motif pruning, and AUCell scoring—will be executed blazingly fast in the Python backend.

```{r grn-run, eval=FALSE}
sce <- RunSCENIC(
  sce,
  tfs_path = tfs_path,
  motif_annotations_path = motif_annotations_path,
  database_paths = database_paths,
  assay_use = "counts",
  num_workers = 8,
  name = "scenic_main"
) # Supports multi-threading
```

## Result Retrieval and State Awareness

After the run, where does the massive AUCell scoring matrix (tens of thousands of cells by hundreds of regulons) go?

To keep the main SCE object clean and performant, `sclet` mounts this large scoring matrix as an `altExp` (Alternative Experiment) and records it in the `scenic` state. This means you can retrieve either the active SCENIC run or a specific run by `id`, instead of manually digging through `metadata()`.

```{r grn-result, eval=FALSE}
# Check the currently active SCENIC state provenance
get_scenic(sce)

# Retrieve a specific SCENIC run by id
get_scenic(sce, id = "scenic_main")

# Extract the AUCell matrix for downstream analysis
auc_sce <- SingleCellExperiment::altExp(sce, "SCENIC_AUC")
auc_matrix <- SummarizedExperiment::assay(auc_sce, "AUC")

# Inspect available regulons
head(rownames(auc_matrix))
```

## Visualizing Regulon Activity

Once the AUC matrix is stored, the next question is not "where is the result?", but "can I directly visualize a regulon on my embedding?" That is exactly what `plot_scenic_activity()` is for.

```{r grn-plot, eval=FALSE}
# Make sure you already have a reduction such as UMAP
DefaultReduction(sce) <- "UMAP"

# Plot a single regulon on the active embedding
plot_scenic_activity(
  sce,
  regulon = "STAT1(+)",
  id = "scenic_main"
)
```

**Bottom Line:**
You can now treat SCENIC as a proper state-aware module: run it with a named record, retrieve it via `get_scenic(id = ...)`, and visualize a regulon directly with `plot_scenic_activity()`. From there, you can further perform dimensionality reduction and clustering on the `auc_matrix` to see whether cells group together based on "regulatory network activity" rather than only "gene expression levels".
