# Interactive Data Exploration {#interactive}

Not every question is easiest to answer from a static figure or a scripted summary. Sometimes the fastest way to understand a dataset is to click, filter, inspect, and compare interactively.

`sclet` integrates `iSEE` to provide a Shiny-based interactive explorer for your `SingleCellExperiment` objects.

The examples in this chapter are shown but not executed during book rendering, because they launch an interactive Shiny session rather than producing a static result that can be embedded cleanly in the book. In other words, the issue here is not package availability but session lifetime: a documentation build should finish, whereas an interactive explorer is expected to stay open until the user closes it.

## Launch Explorer

You can launch the explorer directly from your R console:

To reproduce this locally, the only real prerequisites are an `SCE` object with at least one usable assay and dimensional reduction, plus the `iSEE` package.

```{r, eval=FALSE}
library(sclet)
sce <- qs::qread("data/pbmc-3k-sce.qs")

# Launch the app
RunExplorer(sce)
```

This will open a web browser where you can:

- Visualize reduced dimensions (UMAP, PCA)
- Explore gene expression on the plot
- Check column data (QC metrics, Clusters)
- Select points and inspect data

## Customization

`RunExplorer` passes additional arguments to `iSEE::iSEE()`, allowing you to pre-configure panels.

```{r, eval=FALSE}
# Example: Start with just a ReducedDimensionPlot and a FeaturePlot
library(iSEE)
initial <- list(
    ReducedDimensionPlot(PanelId=1L),
    FeatureAssayPlot(PanelId=1L)
)

RunExplorer(sce, initial = initial)
```
