# Milo: differential abundance

`sclet` provides a lightweight wrapper around `miloR` for cell neighbourhood differential abundance (DA) analysis.
The Milo functionality in `sclet` is an optional dependency: you need to install `miloR` to run the examples.

## Requirements

Before running Milo DA, prepare a `SingleCellExperiment` with:

- `reducedDims(sce)$PCA`
- `colData(sce)` containing a sample column and a condition column

UMAP is optional (used for visualization only).

## Two-group DA (GLM)

```{r milo-glm, eval = requireNamespace("miloR", quietly = TRUE) && requireNamespace("MouseThymusAgeing", quietly = TRUE), cache=FALSE}
library(MouseThymusAgeing)
library(scater)

sce <- MouseSMARTseqData()
colData(sce)$Sample <- paste(colData(sce)$SortDay, colData(sce)$Age, sep = "_")

sce <- logNormCounts(sce)
sce <- runPCA(sce)
sce <- runUMAP(sce)

res <- runMilo(
  sce = sce,
  sample_col = "Sample",
  condition_col = "Age",
  contrasts = c("Age1wk - Age4wk"),
  k1 = 11, k2 = 11, d1 = 40, d2 = 40, prop = 0.2
)

da <- res$da_results
table(da$SpatialFDR < 0.1)
head(da)
```

## Multi-contrast testing

Use `per_contrast = FALSE` to fit one model and test multiple contrasts jointly. Use `per_contrast = TRUE` to run each contrast and stack results.

```{r milo-multi-contrast, eval = requireNamespace("miloR", quietly = TRUE) && requireNamespace("MouseThymusAgeing", quietly = TRUE), cache=FALSE}
contrasts <- c(
  "Age1wk - Age4wk",
  "Age4wk - Age16wk",
  "Age16wk - Age32wk",
  "Age32wk - Age52wk"
)

res2 <- runMilo(
  sce = sce,
  sample_col = "Sample",
  condition_col = "Age",
  contrasts = contrasts,
  per_contrast = TRUE,
  k1 = 11, k2 = 11, d1 = 40, d2 = 40, d3 = 40,
  prop = 0.2,
  refinement_scheme = "graph"
)

head(res2$da_results)
```

## GLMM (random intercept)

When samples are not independent (e.g., repeated measures), a random intercept can help control correlation structure.

```{r milo-glmm, eval = requireNamespace("miloR", quietly = TRUE) && requireNamespace("scRNAseq", quietly = TRUE), cache=FALSE}
library(scRNAseq)
library(scater)
library(scran)
library(scuttle)

sce2 <- KotliarovPBMCData(mode = "rna", ensembl = TRUE)
colData(sce2)$ObsID <- paste(colData(sce2)$tenx_lane, colData(sce2)$sampleid, sep = "_")

set.seed(42)
keep_samps <- sample(unique(colData(sce2)$ObsID), size = 40)
sce2 <- sce2[, colData(sce2)$ObsID %in% keep_samps]

sce2 <- logNormCounts(sce2)
sce2 <- runPCA(sce2, ncomponents = 30)
sce2 <- runUMAP(sce2, pca = 30)

res3 <- runMilo(
  sce = sce2,
  sample_col = "ObsID",
  condition_col = "adjmfc.time",
  contrasts = c("adjmfc.timed0 high"),
  glmm = TRUE,
  random_effect = "tenx_lane",
  k1 = 60, k2 = 60, d1 = 30, d2 = 30, d3 = 30,
  prop = 0.05,
  overlap = 25
)

head(res3$da_results)
```

## Refit after dropping separated neighbourhoods

In GLMM mode, some neighbourhoods can show perfect separation and return unstable estimates. You can diagnose with `miloR::checkSeparation()` and refit after dropping those neighbourhoods.

```{r milo-refit, eval = requireNamespace("miloR", quietly = TRUE) && exists("res3")}
design_df <- res3$sce@metadata$milo_results$design_df

bad_hood <- which(miloR::checkSeparation(
  res3$milo,
  design.df = design_df,
  condition = "adjmfc.time",
  min.val = 10
))

refit <- refit_milo(
  milo = res3$milo,
  design_df = design_df,
  bad_hood = bad_hood,
  formula = res3$sce@metadata$milo_results$formula,
  contrasts = res3$sce@metadata$milo_results$contrasts,
  glmm = TRUE,
  old_results = res3$da_results
)

head(refit$da_clean)
```
