# Milo: Differential Abundance

Changes in gene expression are only part of the story. In many studies, the more important biological signal is that certain cell states become more or less abundant across conditions, donors, or time points.

`sclet` provides a state-aware interface to `miloR` for cell neighbourhood differential abundance (DA) analysis.
The Milo functionality in `sclet` is an optional dependency: you need to install `miloR` to run the examples. In the GitHub Actions bookdown workflow, these chunks are executed when the required packages are available, so this chapter also acts as a regression check for the DA integration layer.

## Requirements

Before running Milo DA, prepare a `SingleCellExperiment` with:

- `reducedDims(sce)$PCA`
- `colData(sce)` containing a sample column and a condition column

UMAP is optional (used for visualization only).

## Two-group DA (GLM)

```{r milo-glm, eval = eval_chunks && requireNamespace("miloR", quietly = TRUE) && requireNamespace("MouseThymusAgeing", quietly = TRUE), cache=FALSE}
library(MouseThymusAgeing)
library(scater)
library(sclet)

sce <- MouseSMARTseqData()
colData(sce)$Sample <- paste(colData(sce)$SortDay, colData(sce)$Age, sep = "_")

sce <- NormalizeData(sce)
sce <- RunPCA(sce)
sce <- RunUMAP(sce)

res <- RunMilo(
  sce = sce,
  sample_col = "Sample",
  condition_col = "Age",
  name = "milo_main",
  contrasts = c("Age1wk - Age4wk"),
  k1 = 11, k2 = 11, d1 = 40, d2 = 40, prop = 0.2
)

da <- res$da_results
table(da$SpatialFDR < 0.1)
head(da)
get_milo(res$sce)
```

## Multi-contrast testing

Use `per_contrast = FALSE` to fit one model and test multiple contrasts jointly. Use `per_contrast = TRUE` to run each contrast and stack results.

```{r milo-multi-contrast, eval = eval_chunks && requireNamespace("miloR", quietly = TRUE) && requireNamespace("MouseThymusAgeing", quietly = TRUE), cache=FALSE}
contrasts <- c(
  "Age1wk - Age4wk",
  "Age4wk - Age16wk",
  "Age16wk - Age32wk",
  "Age32wk - Age52wk"
)

res2 <- RunMilo(
  sce = sce,
  sample_col = "Sample",
  condition_col = "Age",
  name = "milo_multi",
  contrasts = contrasts,
  per_contrast = TRUE,
  k1 = 11, k2 = 11, d1 = 40, d2 = 40, d3 = 40,
  prop = 0.2,
  refinement_scheme = "graph"
)

head(res2$da_results)
```

## GLMM (random intercept)

When samples are not independent (e.g., repeated measures), a random intercept can help control correlation structure.

```{r milo-glmm, eval = eval_chunks && requireNamespace("miloR", quietly = TRUE) && requireNamespace("scRNAseq", quietly = TRUE) && requireNamespace("ensembldb", quietly = TRUE) && requireNamespace("AnnotationHub", quietly = TRUE), cache=FALSE}
library(scRNAseq)
library(scater)
library(scran)
library(scuttle)
library(sclet)

sce2 <- KotliarovPBMCData(mode = "rna", ensembl = FALSE)
colData(sce2)$ObsID <- paste(colData(sce2)$tenx_lane, colData(sce2)$sampleid, sep = "_")

set.seed(42)
keep_samps <- sample(unique(colData(sce2)$ObsID), size = 40)
sce2 <- sce2[, colData(sce2)$ObsID %in% keep_samps]

sce2 <- NormalizeData(sce2)
sce2 <- RunPCA(sce2, ncomponents = 30)
sce2 <- RunUMAP(sce2, dims = 1:30)

res3 <- RunMilo(
  sce = sce2,
  sample_col = "ObsID",
  condition_col = "adjmfc.time",
  name = "milo_glmm",
  contrasts = c("adjmfc.timed0 high"),
  glmm = TRUE,
  random_effect = "tenx_lane",
  k1 = 60, k2 = 60, d1 = 30, d2 = 30, d3 = 30,
  prop = 0.05,
  overlap = 25
)

head(res3$da_results)
```

## Refit after dropping separated neighbourhoods

In GLMM mode, some neighbourhoods can show perfect separation and return unstable estimates. You can diagnose with `miloR::checkSeparation()` and refit after dropping those neighbourhoods.

If you keep multiple Milo runs in the same object, give them different `name` values and inspect a specific record with `get_milo(sce, id = "your_id")`.

```{r milo-refit, eval = eval_chunks && requireNamespace("miloR", quietly = TRUE) && exists("res3")}
milo_info <- get_milo(res3$sce)
design_df <- milo_info$design_df

bad_hood <- which(miloR::checkSeparation(
  res3$milo,
  design.df = design_df,
  condition = "adjmfc.time",
  min.val = 10
))

refit <- refit_milo(
  milo = res3$milo,
  design_df = design_df,
  bad_hood = bad_hood,
  formula = milo_info$formula,
  contrasts = milo_info$contrasts,
  glmm = TRUE,
  old_results = res3$da_results
)

head(refit$da_clean)
```
