# Multimodal Expansion {#multimodal-expansion}

This chapter is the reserved mainline for ADT, ATAC, and multiome workflows. The object contract is now in place so that modality-specific analysis backends can be plugged in later without breaking the navigation model.

```{r multimodal-setup, include=FALSE}
# NOTE: Chunks use eval=FALSE because the PBMC demo lacks ADT/ATAC altExps.
# has_multimodal_adt/atac can run in CI (returns FALSE on the demo object).
# RunMultimodalWorkflow returns early with a message when no altExps are found.
```

## What Is Already Available

`sclet` now provides lightweight modality helpers and a workflow shell:

- `has_multimodal_adt(sce)` and `has_multimodal_atac(sce)` check whether `altExp(sce, "ADT")` or `altExp(sce, "ATAC")` is present.
- `register_multimodal_adt(sce)` and `register_multimodal_atac(sce)` record the modality metadata in the typed state layer, including feature and cell dimensions.
- `RunMultimodalWorkflow()` auto-detects available altExps and registers them as a coherent multimodal analysis mainline record.

The intent is not to promise a large backend surface too early, but to make the book structure honest and to establish the contract that future modality analysis will extend.

## Detecting and Registering Modalities

The helpers are designed to be used before and after loading multimodal data into the object.

```{r multimodal-register, eval=FALSE}
library(sclet)

# After loading data, check which modalities are present
has_multimodal_adt(sce)   # TRUE if altExp(sce, "ADT") exists
has_multimodal_atac(sce)  # TRUE if altExp(sce, "ATAC") exists

# Register modalities in the state layer
if (has_multimodal_adt(sce)) {
    sce <- register_multimodal_adt(sce, name = "my_ADT")
}

if (has_multimodal_atac(sce)) {
    sce <- register_multimodal_atac(sce, name = "my_ATAC")
}
```

## One-Step Registration

For convenience, `RunMultimodalWorkflow()` detects and registers all available modalities in one call. When no altExps are found, it returns the object unchanged with a message.

```{r multimodal-workflow, eval=FALSE}
sce <- RunMultimodalWorkflow(sce, name = "my_multimodal")
```

## Planned Interface Direction

The current design notes point to three priorities for future work:

- **altExp semantics**: `altExp("ADT")` and `altExp("ATAC")` are the canonical modality accessors, and the registration helpers are the bridge between the Bioconductor object model and `sclet`'s state contract.
- **State meaning**: each modality gets its own `multimodal_*` record in the analysis layer and its own entry under the `multimodal` state type, so downstream code can inspect the modality landscape without guessing.
- **Analysis backends**: once the object contract is stable, the next step is to wire modality-specific analysis functions (e.g., ADT-based clustering, ATAC peak calling) into the same `Run*` / `get_*` / `has_*` pattern used throughout `sclet`.

## Current Guidance

Until dedicated multimodal analysis backends are implemented:

- Use `sclet`'s single-cell mainlines as the stable baseline for RNA analysis.
- Attach ADT or ATAC data through `altExp(sce, "ADT") <- adt_sce` and register it with `register_multimodal_adt(sce)`.
- Keep modality-specific analysis explicit in your own scripts for now; the object contract ensures that when backends arrive, the transition will be a matter of calling a `Run*` function on an already-registered modality.
