# Automation & Provenance Logs {#pipeline-logs}

Ever found yourself lost in parameter hell? You spend a week tweaking PCA dimensions, resolution parameters, and variable feature counts. Three months later, when it's time to write the "Materials and Methods" section for your manuscript, you stare at your `SingleCellExperiment` object and think: *"Wait, did I use 2000 or 3000 highly variable genes for this?"*

`sclet` solves this pain point beautifully with two features: **Automated Pipelines** and **Provenance Logs**.

In the bookdown workflow, the examples below are executed against a packaged PBMC object rather than a user-specific raw directory. This keeps the chapter reproducible while still showing the intended day-to-day workflow.

```{r pipeline-setup, include=FALSE, eval=TRUE}
pipeline_ready <- eval_chunks

if (pipeline_ready) {
    library(sclet)
    sce_pipeline_input <- readRDS("data/pbmc-3k-sce.rds")
    sce_pipeline_input <- sce_pipeline_input[, seq_len(min(800, ncol(sce_pipeline_input)))]
}
```

## The One-Liner Pipeline

If you just want to get from a raw count matrix to a UMAP plot as fast as possible without writing 10 lines of boilerplate code, `RunStandardPipeline()` is your best friend.

```{r pipeline, eval=pipeline_ready}
# In a real project you might start from:
# sce <- Read10X(data.dir = "filtered_feature_bc_matrix/")
#
# For a runnable documentation example we reuse a packaged PBMC object.
sce <- sce_pipeline_input

# Run everything from normalization to UMAP in one line
sce <- RunStandardPipeline(
    sce, 
    nfeatures = 2000, 
    dims = 1:20, 
    resolution = 0.8
)

# Boom. Ready to inspect.
sce
```

Under the hood, `RunStandardPipeline()` sequentially executes:
1. `NormalizeData()`
2. `FindVariableFeatures()`
3. `ScaleData()`
4. `RunPCA()`
5. `FindNeighbors()`
6. `FindClusters()`
7. `RunUMAP()`

## The Provenance Log (CommandLog)

The true magic happens *after* you run your analysis. Every major function in `sclet` (whether you use the pipeline or run them step-by-step) automatically records its exact parameters and timestamp into the object's `CommandLog`.

You can view the raw log anytime:

```{r cmdlog, eval=pipeline_ready}
# Peek at the raw command logs
CommandLog(sce, details = TRUE)
```

## Writing Methods for You: `PipelineSummary`

Looking at a raw list of parameters is better than nothing, but it's not exactly manuscript-ready. This is where `PipelineSummary()` shines. It reads the `CommandLog` and translates it into a human-readable English summary.

```{r summary, eval=pipeline_ready}
# Generate a manuscript-ready methods summary
PipelineSummary(sce)
```

**Output example:**
> The data was normalized using the LogNormalize method. Highly variable features were identified using the vst method (n=2000). The data was scaled and centered. Principal Component Analysis (PCA) was performed, and the first 20 dimensions were used for downstream analysis. A K-nearest neighbor (KNN) graph was constructed based on these dimensions (k.param=20). Cells were clustered using the Louvain algorithm with a resolution of 0.8. Finally, Uniform Manifold Approximation and Projection (UMAP) was run for 2D visualization.

Life is simply too short to manually write boilerplate methods sections. Let `sclet` do the heavy lifting for you.
