# Program, Regulon, and Mechanistic Interpretation {#program-regulon}

Your clustering is done. Your trajectory is drawn. You have a list of marker genes. But the most important question remains unanswered: **why** are those states different? Which pathways drive the identity of each cluster? Which transcription factors control the developmental transitions? And can you summarize dozens of marker genes into a handful of interpretable "programs" that a biologist can reason about?

This mainline answers that downstream question. In `sclet`, signature scoring, pathway-level summarization, regulon inference, and program access are not scattered tools—they form one interpretation chain that moves from raw gene lists to mechanistically meaningful summaries.

```{r program-setup, include=FALSE}
# NOTE: RunKEGG/RunSCENIC/RunGRN chunks use eval=FALSE because:
# - pySCENIC needs Python basilisk + large cisTarget feather DB (~297 MB)
# - Small ref files (TFs list + motif annotations) are in data-raw/scenic-refs.R
# RunGeneSetScoring and plot_* chunks using the PBMC object run in CI.
```

The high-level direction is to let users move from quick activity scoring to richer mechanism-oriented models without changing the object contract. `RunProgramWorkflow()` is the intended mainline entry; the chapters below document the atomic building blocks that already make up that workflow.

## Gene set scoring and pathway activity

Often, relying on the expression of a single gene (like a traditional marker) introduces too much noise and dropout artifacts. If you want to know whether a cell population is undergoing "Hypoxia" or has an active "Cell Cycle", scoring the entire pathway or gene set is the most robust approach.

However, the algorithm landscape is fragmented: `AUCell`, `UCell`, `GSVA`... Each backend has a different output format. If you want to plot the results downstream, standardizing these formats can be a nightmare.

This is exactly the kind of problem `sclet` is designed to simplify. We provide the `RunGeneSetScoring()` interface, which unifies these backends and injects the scores directly into the state-aware workflow and `colData`, so the results remain easy to inspect and easy to reuse later.

## Multi-Backend Scoring in One Click

Let's define a few gene sets of interest and run the scoring on our PBMC dataset.

```{r geneset-scoring-setup}
library(sclet)

# Define signatures
my_signatures <- list(
  T_cell = c("CD3D", "CD3E", "CD8A", "CD4"),
  B_cell = c("CD79A", "MS4A1"),
  Myeloid = c("LYZ", "CD14", "FCGR3A")
)
```

You can decisively choose the backend that best fits your needs. `UCell` is rank-based, extremely fast, and highly suitable for single-cell data.

```{r geneset-scoring-run}
# Use UCell backend
pbmc <- RunGeneSetScoring(pbmc, gene_sets = my_signatures, method = "UCell", name = "UCellScore")
```

## Program mainline workflow

If you want to organize pathway scoring and regulon activity as one coherent interpretation mainline, use `RunProgramWorkflow()` as the semantic entry point. It keeps the analysis story centered on "which programs are active" instead of "which backend did I call".

```{r geneset-workflow}
pbmc_program <- pbmc
pbmc_program <- RunProgramWorkflow(
  pbmc_program,
  gene_sets = my_signatures,
  scoring_method = "UCell",
  name = "program_main"
)
```

## Reading and Visualizing Results

Where do the scores go after calculation?
1. By default, they are added as columns in the cellular metadata (`colData`).
2. The action is recorded in the `geneset_scoring` state machine.

Because the scores are now registered as a state-aware analysis result, we can inspect their provenance and directly plot a single program through a unified activity consumer:

```{r geneset-plot, fig.width=12, fig.height=5}
# Inspect the scoring state provenance
get_geneset_scoring(pbmc, id = "UCellScore")

# Visualize the T_cell score distribution on the UMAP space
plot_program_activity(pbmc, program = "T_cell", source = "geneset_scoring", id = "UCellScore")

# Summarize multiple programs across groups
plot_program_heatmap(
  pbmc,
  programs = c("T_cell", "B_cell", "Myeloid"),
  source = "geneset_scoring",
  group.by = "colLabels",
  id = "UCellScore"
)
```

```{r geneset-workflow-plot, fig.width=12, fig.height=5}
plot_program_activity(
  pbmc_program,
  program = "T_cell",
  source = "geneset_scoring",
  id = "program_main_scores"
)

plot_program_heatmap(
  pbmc_program,
  programs = c("T_cell", "B_cell", "Myeloid"),
  source = "geneset_scoring",
  group.by = "colLabels",
  id = "program_main_scores"
)
```

## Advanced Pathway Discovery (altExp and RunKEGG)

When dealing with hundreds of pathways (like the entire KEGG database), dumping all scores into `colData` is messy. Instead, `sclet` allows storing these massive score matrices as an alternative experiment (`altExp`). This keeps your metadata clean while retaining perfect alignment with your cells.

Furthermore, `sclet` provides `RunKEGG()` as a one-stop wrapper. It automatically downloads pathways via `clusterProfiler`, scores the cells using `UCell`, stores the matrix in an `altExp`, and even identifies cell-type specific marker pathways using robust Wilcoxon rank-sum tests via `FindMarkerPathways()`.

```{r kegg-scoring, eval=FALSE}
# Run the complete KEGG workflow
# Note: Requires org.Hs.eg.db for human data
res <- RunKEGG(pbmc, species = "hsa", method = "UCell", group.by = "seurat_annotations")

# The updated SCE object with the KEGG altExp
pbmc <- res$sce

# The marker pathways data.frame
kegg_markers <- res$markers
head(kegg_markers)
```

You can also manually run differential testing on any pathway scores stored in an `altExp`:

```{r marker-pathways, eval=FALSE}
# If you ran RunGeneSetScoring with as_altExp = TRUE and name = "MyScores"
markers <- FindMarkerPathways(pbmc, altExp_name = "MyScores", group.by = "seurat_annotations")
```

**Bottom Line:**
Regardless of whether the underlying engine is `UCell` or `AUCell`, you can immediately plot the results with `FeatureDimPlot` or seamlessly run differential analysis. Life is too short to wrangle metadata formats.

## SCENIC regulon inference

How do Transcription Factors (TFs) regulate target genes? Traditional differential expression analysis only tells you "which genes changed," but SCENIC tells you "who is issuing the commands."

If you have ever run the legacy R version of SCENIC, you undoubtedly remember its excruciatingly slow speed and the requirement to download tens of gigabytes of local databases. Today, the Python-based `pySCENIC` (powered by GRNBoost2) completely crushes the original R version in both speed and performance.

Rather than forcing users to leave the `SingleCellExperiment` workflow and manually manage a separate Python environment, `sclet` brings this functionality back into the same state-aware framework. Leveraging the `basilisk` isolated environment, it provides a native `sclet` entry point to `pySCENIC`: `RunSCENIC()`.

The examples in this chapter are shown but not executed during book rendering. A realistic `pySCENIC` workflow depends on external transcription-factor lists and cisTarget ranking databases, so it is more honest to present the code as a template that readers can adapt to their own resources. In practice, the limiting factor is not the Python environment itself, but the requirement to provide large database files that are version-matched to the species and genome build of your dataset.

## Environment and Data Preparation

You no longer need to manually configure complex Python environments or wrestle with `conda` conflicts. `sclet` securely prepares the sandbox in the background. You only need to provide three things:
1. Your single-cell expression matrix (SCE object)
2. A list of transcription factors (`.txt`)
3. The Motif database file (in `Feather` format)

For most human or mouse analyses, this also means checking that your TF list, motif annotation table, and cisTarget ranking database all come from the same release family before you start the run.

```{r grn-setup, eval=FALSE}
library(sclet)
# Assuming 'sce' is your pre-processed object

# Specify the paths to your downloaded databases
tfs_path <- "path/to/hs_hgnc_tfs.txt"
database_paths <- c(
  "path/to/hg38__refseq-r80__10kb_up_and_down_tss.mc9nr.feather"
)
motif_annotations_path <- "path/to/motifs-v9-nr.hgnc-m0.001-o0.0.tbl"
```

## One-Click Inference and Scoring

Next, call `RunSCENIC()`. All heavy lifting—GRNBoost2 inference, Motif pruning, and AUCell scoring—will be executed blazingly fast in the Python backend.

```{r grn-run, eval=FALSE}
sce <- RunSCENIC(
  sce,
  tfs_path = tfs_path,
  motif_annotations_path = motif_annotations_path,
  database_paths = database_paths,
  assay_use = "counts",
  num_workers = 8,
  name = "scenic_main"
) # Supports multi-threading
```

## Result Retrieval and State Awareness

After the run, where does the massive AUCell scoring matrix (tens of thousands of cells by hundreds of regulons) go?

To keep the main SCE object clean and performant, `sclet` mounts this large scoring matrix as an `altExp` (Alternative Experiment) and records it in the `scenic` state. This means you can retrieve either the active SCENIC run or a specific run by `id`, instead of manually digging through `metadata()`.

```{r grn-result, eval=FALSE}
# Check the currently active SCENIC state provenance
get_scenic(sce)

# Retrieve a specific SCENIC run by id
get_scenic(sce, id = "scenic_main")

# Extract the AUCell matrix for downstream analysis
auc_sce <- SingleCellExperiment::altExp(sce, "SCENIC_AUC")
auc_matrix <- SummarizedExperiment::assay(auc_sce, "AUC")

# Inspect available regulons
head(rownames(auc_matrix))
```

## Visualizing Regulon Activity

Once the AUC matrix is stored, the next question is not "where is the result?", but "can I directly visualize a regulon on my embedding?" That is exactly what `plot_scenic_activity()` is for.

```{r grn-plot, eval=FALSE}
# Make sure you already have a reduction such as UMAP
DefaultReduction(sce) <- "UMAP"

# Plot a single regulon on the active embedding
plot_scenic_activity(
  sce,
  regulon = "STAT1(+)",
  id = "scenic_main"
)
```

**Bottom Line:**
You can now treat SCENIC as a proper state-aware module: run it with a named record, retrieve it via `get_scenic(id = ...)`, and visualize a regulon directly with `plot_scenic_activity()`. From there, you can further perform dimensionality reduction and clustering on the `auc_matrix` to see whether cells group together based on "regulatory network activity" rather than only "gene expression levels".

## Unified GRN Entry: `RunGRN()`

If your question is simply "what gene regulatory networks are active in my cells?", the implementation detail of which backend you used should not matter. `RunGRN()` wraps `RunSCENIC()` as the default backend while leaving room for future methods (decoupleR, FigR, SCENIC+) through the same interface.

```{r grn-unified, eval=FALSE}
# Same parameters as RunSCENIC, different semantic framing
sce <- RunGRN(
    sce,
    tfs_path = "allTFs.txt",
    motif_annotations_path = "motifs.tbl",
    database_paths = c("hg38_10kbp.feather"),
    name = "grn_main"
)
```

## Unified Program Access: `get_program()` / `has_program()`

Once you have signature scores and regulon activities in your object, you should not need to remember which source they came from. `get_program()` transparently resolves a program name from either gene set scoring or SCENIC, returning a per-cell activity vector. `has_program()` is the corresponding boolean guard.

```{r program-accessor, eval=FALSE}
# Check if a program exists
has_program(sce, "T_cell")

# Get activity vector
t_activity <- get_program(sce, "T_cell", source = "auto")
head(t_activity)

# Same interface works for regulons
has_program(sce, "STAT1(+)", source = "scenic")
stat1 <- get_program(sce, "STAT1(+)", source = "scenic")
```

## Program Activity Dotplot: `plot_program_dotplot()`

When you want a single overview of multiple programs across cell groups, `plot_program_dotplot()` summarizes each program per group with dot size reflecting the fraction of expressing cells and color reflecting mean activity. This is the program-level equivalent of Seurat's DotPlot, consuming the same typed state records as `get_program()`.

```{r program-dotplot, eval=FALSE}
plot_program_dotplot(
    sce,
    programs = c("T_cell", "B_cell", "Myeloid", "STAT1(+)"),
    source = "auto",
    group.by = "colLabels"
)
```
