# Cell Type Annotation

After clustering, the natural next question is simple: what do these clusters actually represent? In practice, annotation usually combines two strategies. We first inspect clusters manually using known markers, and then we use a reference-based method to make the process faster, more systematic, and easier to reproduce.

This chapter introduces both perspectives. It keeps the biological intuition of manual annotation, while emphasizing the state-aware analysis interface that makes annotation results reusable in later steps.

## Manual annotation

Please refer to the [Cell cluster annotation](#manual-annotation) session. 


## Automatic annotation
    
The low-level `SingleR::SingleR()` call is still useful for understanding the underlying method. In daily `sclet` workflows, however, the recommended entry point is `RunSingleR()`, because it records the annotation result and its provenance back into the unified state layer.

Alternatively, `sclet` also provides a lightweight reference mapping implementation via `RunKNNPredict()` (or `RunReferenceMapping()`), which transfers labels from a reference dataset to your query dataset based on KNN.

In this chapter, both the low-level `SingleR()` workflow and the `sclet` analysis examples are executed during book rendering when the required Bioconductor dependencies are available. This keeps the chapter useful as documentation while also exercising the annotation layer in CI.

### Lightweight Reference Mapping (KNN)

If you have a reference `SingleCellExperiment` with cell type annotations, you can use it to map labels to your query dataset quickly:

```{r run-knn-predict, eval = eval_chunks && requireNamespace("BiocNeighbors", quietly = TRUE)}
set.seed(1)
ref_idx <- sample(seq_len(ncol(pbmc)), 500)
ref_sce <- pbmc[, ref_idx]
query_sce <- pbmc[, -ref_idx]

ref_sce$label <- as.character(SingleCellExperiment::colLabels(ref_sce))

query_sce <- RunKNNPredict(
    object = query_sce,
    ref = ref_sce,
    labels = "label",
    layer = "logcounts",
    k = 5,
    name = "knn_demo"
)

# Inspect the recorded reference-mapping state
get_mapping(query_sce)

# Visualize the query and reference in the same projection space
ProjectionPlot(query_sce, ref_sce)
```

### Run SingleR with sclet

```{r run-singleR-sclet, eval = eval_chunks && requireNamespace("SingleR", quietly = TRUE) && requireNamespace("celldex", quietly = TRUE)}
library(sclet)
hpca <- celldex::HumanPrimaryCellAtlasData()

pbmc2 <- RunSingleR(pbmc, ref = hpca, labels = hpca$label.main)

# explicitly choose the layer used for annotation
pbmc2 <- RunSingleR(pbmc, ref = hpca, labels = hpca$label.main, layer = "logcounts")

# inspect the recorded annotation state
get_annotation(pbmc2)

# inspect the recorded reference-mapping state
get_mapping(pbmc2)
```


### Run SingleR

```{r run-singleR, eval = eval_chunks && requireNamespace("SingleR", quietly = TRUE) && requireNamespace("celldex", quietly = TRUE)}

hpca <- celldex::HumanPrimaryCellAtlasData()
hpca
res <- SingleR(test = assay(pbmc, "logcounts"), ref = hpca, labels = hpca$label.main)
head(res)


dice <- celldex::DatabaseImmuneCellExpressionData()
res2 <- SingleR(test = assay(pbmc, "logcounts"), ref = dice, labels = dice$label.main)
head(res2)
```

When you use the low-level `SingleR()` call directly, you are responsible for deciding which matrix to pass. In contrast, `RunSingleR()` follows the current `sclet` layer contract and resolves that choice from `DefaultLayer(pbmc)` unless you override it.

If the current layer comes from a corrected integration workflow, both the annotation state and the mapping state record that provenance. In practice, this means `get_annotation(pbmc)$inputs$integration` and `get_mapping(pbmc)$inputs$integration` can tell you whether the label transfer was performed on top of a corrected representation rather than the raw normalized layer.

### Annotate with SingleR result

```{r annotation, eval = eval_chunks && exists("res") && exists("res2")}
pbmc[['hpca_label']] <- res$labels

pbmc[['dice_label']] <- res2$labels
```


### Comparision with manual annotation result

```{r annotation-comparison, fig.width=16, fig.height=6, eval = eval_chunks && exists("res") && exists("res2")}
# Ensure manual labels are available in colData for comparison
pbmc$label <- SingleCellExperiment::colLabels(pbmc)

x <- colData(pbmc)[, c("label", "hpca_label", "dice_label")]

plot_list(
    manual = sc_dim(pbmc, reduction="UMAP"),    
    hpca = sc_dim(pbmc, reduction="UMAP", mapping=aes(color=hpca_label)),
    dice = sc_dim(pbmc, reduction="UMAP", mapping=aes(color=dice_label)),
    ncol = 3) & 
sc_dim_geom_label(geom = ggrepel::geom_text_repel) & 
theme(legend.position='inside')

table(x[,c(1,2)])

table(x[,c(1,3)])
```

