# Spatial Context and Niche Analysis {#spatial-context}

This mainline extends the analysis map from dissociated single cells back into tissue context. The user question is no longer only which states exist, but where they sit, which compositions change across space, and how niche structure constrains interpretation.

`sclet` now provides `RunSpatialWorkflow()` as the semantic shell, with `RunSpatialDeconvolution()` as the primary backend and `RunSpatialColocalization()` and `RunSpatialNiche()` reserved for future integration. All spatial functions now follow the same state contract as other `sclet` modules: each runs records a `name` identifier, writes into the typed state record layer, and logs the command for downstream provenance tracking.

## Spatial Mainline Workflow

The simplest entry point is the workflow shell. It dispatches deconvolution through cell2location and reserves colocalization and niche slots.

```{r spatial-workflow, eval=FALSE}
library(sclet)
# sce_spatial: spatial SingleCellExperiment with 'counts' assay
# sce_ref: reference scRNA-seq SCE with cell type labels

sce_spatial <- RunSpatialWorkflow(
    sce_spatial,
    sce_ref = sce_ref,
    ref_group_key = "cell_type",
    run_deconvolution = TRUE,
    name = "my_spatial_analysis"
)

# Check the active spatial state
has_spatial(sce_spatial)
get_spatial(sce_spatial)
```

## High-Resolution Spatial Deconvolution: Cell2location

`cell2location` is widely recognized as the "gold standard" for spatial deconvolution. It uses a negative binomial regression model to map scRNA-seq reference data onto spatial spots, inferring the true absolute abundance of each cell type within a given location.

In `sclet`, this complex Bayesian inference is simplified into a single line of code:

To run this locally, the minimum prerequisites are a spatial object with a `counts` assay, a reference `SCE` with the same gene space, and a biologically meaningful cell-type column in `colData(sce_ref)`. The `name` parameter lets you tag each run so that multiple deconvolution results can coexist in the same object.

```{r spatial-deconv, eval=FALSE}
library(sclet)
# Assume 'sce_spatial' is your spatial transcriptomics object, 
# and 'sce_ref' is your single-cell reference object with a "cell_type" column.

sce_spatial <- RunSpatialDeconvolution(
  sce_spatial, 
  sce_ref, 
  ref_group_key = "cell_type",
  max_epochs = 50,
  name = "my_deconv"
)
```

**Under the Hood:**
The function invokes the GPU (if available) via the `basilisk` environment to train the model. Once complete, the predicted abundances for each cell type are directly appended to `colData(sce_spatial)`, and the `spatial` state is activated through `sclet_set_analysis_state()`. Each run is also logged via `sclet_log_command()`, so `CommandLog(sce_spatial)` will show the parameters and timestamp. You can immediately use spatial plotting functions to visualize the distribution of specific cell types across the tissue slice.

## Spatial Colocalization (Reserved)

`RunSpatialColocalization()` is a reserved interface for future colocalization backends. Calling it now records the analysis intent as a typed state record, which means downstream code can already use `has_colocalization()` and `get_colocalization()` as guard and accessor, even before a concrete backend is wired in.

```{r spatial-coloc-reserved, eval=FALSE}
sce <- RunSpatialColocalization(sce, name = "coloc_demo")

has_colocalization(sce)

get_colocalization(sce)
```

## Spatial Niche Detection (Reserved)

`RunSpatialNiche()` follows the same pattern: reserved now, ready for a backend later. `has_spatial_niche()` and `get_spatial_niche()` are already exported and can be used in downstream scripts and plotting functions.

```{r spatial-niche-reserved, eval=FALSE}
sce <- RunSpatialNiche(sce, name = "niche_demo")

has_spatial_niche(sce)

get_spatial_niche(sce)
```

## Inspecting Spatial Results Across Layers

Because all spatial functions now register through the unified state layer, a single accessor pattern works for deconvolution, colocalization, and niche results alike. When you keep multiple spatial records, use `id` to address a specific one.

```{r spatial-accessors-pattern, eval=FALSE}
# The default spatial record (active state)
has_spatial(sce)
get_spatial(sce, element = "columns")

# A specific spatial record by id
get_spatial(sce, id = "my_deconv")

# Colocalization and niche records have their own accessors
has_colocalization(sce)
get_colocalization(sce, element = "status")

has_spatial_niche(sce)
```

## In-Silico Gene Perturbation: CellOracle

Differential expression tells you "what changed," SCENIC tells you "who is controlling," but **CellOracle** allows you to conduct **thought experiments (In-silico Perturbation)**.

Imagine asking: *If I knockout the GATA2 gene in hematopoietic stem cells, how will their developmental trajectory shift?* Performing this in vitro is costly and time-consuming. In `sclet`, it takes only a few minutes of simulation.

This simulation only makes sense after you have already constructed a compatible GRN prior and a dimensional representation that captures the state manifold you want to perturb.

```{r spatial-perturb, eval=FALSE}
# Assume 'sce' has already been processed through SCENIC (to obtain a GRN), 
# has an active dimensional reduction space, and you have exported a base GRN file.
base_grn_path <- "path/to/celloracle_base_grn.csv"

sce <- RunInSilicoPerturbation(
  sce, 
  target_gene = "GATA2", 
  perturbation_value = 0.0, # Simulate a full knockout
  base_grn_path = base_grn_path,
  cluster_key = "cell_type",
  reduction = "PCA"
)

# Inspect the perturbation state
get_perturbation(sce)
```

**The Impact:**
The inferred cellular shift vectors are carefully recorded. When combined with streamplots or vector field visualizations, you can intuitively observe on the UMAP space: in a parallel universe without *GATA2*, which terminal state are these cells "fleeing" from, and which incorrect lineage are they "rushing" toward?
