# State-Aware Analysis Contract {#state-aware}

Ever found yourself lost in "parameter hell"?

After running batch correction, you have to memorize whether your reduced dimension matrix is named `fastMNN` or `scVI`. When you move downstream to visualization, you have to manually specify `reduction = "fastMNN"`. If you introduce trajectory inference, gene set scoring, or spatial deconvolution along the way, the scattered results wrap around your `SingleCellExperiment` (SCE) object like spaghetti. Hand this analysis over to a colleague—or return to it yourself three months later—and no one will remember which `umap` was derived from which batch-corrected matrix.

This is a common systemic flaw in transcriptomic workflows: **We have the data, but we lose the context.**

To solve this, `sclet` introduces its core design philosophy: the **Analysis-State Contract**.

## What is the State Contract?

In `sclet`, the data flow adapts to your analysis pipeline, rather than forcing you to memorize key names. When you run `RunIntegration()`, the object remembers the "active integration state". When you subsequently run `RunUMAP()`, it automatically perceives the output from the previous step.

More importantly, `sclet` quietly records a **Directed Acyclic Graph (DAG)** in the background. This is your analysis provenance (or "bloodline").

## Macro View: `Status()`

Let's look at a minimal runnable example. By simply calling `Status()`, `sclet` provides a comprehensive health report of your object:

```{r state-status}
library(sclet)
# Load a pre-processed PBMC dataset
pbmc <- readRDS("data/pbmc-3k-sce.rds")

# Check the current status of the object
Status(pbmc)
```

Whether you are using native R functions or invoking top-tier Python algorithms via the `basilisk` sandbox, the traces they leave are unified and clearly displayed.

## Digging Deeper: `get_analysis_context()`

If you need to inspect the details of a specific state, `sclet` provides a family of `get_*()` and `has_*()` macros.

```{r state-context}
# Get the full DAG provenance tree
# This highly structured context is perfect for feeding into an AI Copilot
ctx <- get_analysis_context(pbmc)

# List all recorded advanced analysis records
names(ctx$records)
```

**Bottom Line:**
Stop wrestling with `reducedDims` and `colData` variable names. Go with the flow, let the `sclet` state machine remember the details, and focus on the actual biology.

## Manual Overrides: Layers and Default APIs

While the state machine handles 99% of routing automatically, there are times when you want manual control. For example, maybe you want to switch the default visualization from UMAP back to t-SNE, or you want to temporarily access the raw, uncorrected counts instead of the integrated layer.

`sclet` provides a suite of intuitive setter/getter macros:

```{r defaults, eval=eval_chunks}
# Check available layers
Layers(pbmc)

# Get the data for a specific layer
raw_data <- LayerData(pbmc, layer = DefaultLayer(pbmc))
dim(raw_data)

# Change the default reduction used by DimPlot, if available
if ("PCA" %in% reducedDimNames(pbmc)) {
    DefaultReduction(pbmc) <- "PCA"
}
DefaultReduction(pbmc)

# Change the default assay to a concrete assay stored in the object
if ("logcounts" %in% assayNames(pbmc)) {
    DefaultAssay(pbmc) <- "logcounts"
}
DefaultAssay(pbmc)
```

These macros mimic the Seurat API perfectly, ensuring that users migrating to `SingleCellExperiment` feel right at home while retaining the robust Bioconductor data structure under the hood.
