# Coarse-graining of large single-cell data into metacells using SuperCell {#supercell}


Here, we use the `pbmc4k` dataset as presented in [Batch correction](https://yulab-smu.top/sclet/batch-correction.html#preprocess). 


## `RunSuperCell` 


The `RunSuperCell()` function is a wrapper function to run [SuperCell](https://github.com/GfellerLab/SuperCell).


```{r run-SC}
sce_sc <- RunSuperCell(pbmc4k, gamma = 5)
sce_sc
```


The output of `RunSuperCell()` is a `SingleCellExperiment` object that stores gene expression matrix of metacells. 

We can use it for downstram analysis.

```{r supercell, fig.width=5, fig.height=4}
post_process <- function(sce) {
    sce <- NormalizeData(sce)
    sce <- FindVariableFeatures(sce)
    sce <- ScaleData(sce)
    sce <- runPCA(sce, subset_row = VariableFeatures(sce), exprs_values = "scaled")
    
    sce <- FindNeighbors(sce, dims = 1:10)
    sce <- FindClusters(sce)
    sce <- RunUMAP(sce)
    return(sce)
}

sce_sc <- post_process(sce_sc)

library(ggsc)
sc_dim(sce_sc, reduction="UMAP") + sc_dim_geom_label()
```


## Estimate SuperCell purity

```{r SC-purity}

pbmc4k2 <- post_process(pbmc4k)

SC <- metadata(sce_sc)$SuperCell

purity <- SuperCell::supercell_purity(pbmc4k2$label, SC$membership, method = 'entropy')

head(purity)

summary(purity)
```
