# Metacells with SuperCell {#supercell}

As single-cell datasets continue to grow, it is often useful to replace highly redundant individual cells with a smaller number of representative metacells. This reduces computational burden, improves robustness in some downstream analyses, and can make large datasets easier to interpret.

In `sclet`, this workflow is exposed through `RunSuperCell()`, which wraps the `SuperCell` package while keeping the result inside a `SingleCellExperiment`-compatible workflow. In this chapter, we use the `pbmc4k` dataset introduced earlier in the batch-correction chapter.

Because `SuperCell` is installed in the bookdown GitHub Actions workflow, the examples in this chapter are executed during documentation build and therefore also serve as regression checks.


## RunSuperCell


The `RunSuperCell()` function provides the main entry point for metacell construction.


```{r run-SC, eval=eval_chunks && requireNamespace("SuperCell", quietly = TRUE)}
sce_sc <- RunSuperCell(pbmc4k, gamma = 5)
sce_sc
```


The output of `RunSuperCell()` is a `SingleCellExperiment` object whose columns now represent metacells rather than individual cells. From this point on, you can continue with a fairly standard downstream workflow.

```{r supercell, fig.width=5, fig.height=4, eval=eval_chunks && requireNamespace("SuperCell", quietly = TRUE)}
post_process <- function(sce) {
    sce <- NormalizeData(sce)
    sce <- FindVariableFeatures(sce)
    sce <- ScaleData(sce)
    sce <- RunPCA(sce, subset_row = VariableFeatures(sce), layer = "scaled")
    
    sce <- FindNeighbors(sce, dims = 1:10)
    sce <- FindClusters(sce)
    sce <- RunUMAP(sce)
    return(sce)
}

sce_sc <- post_process(sce_sc)

library(ggsc)
sc_dim(sce_sc, reduction="UMAP") + sc_dim_geom_label()
```


## Estimate SuperCell Purity

```{r SC-purity, eval=eval_chunks && requireNamespace("SuperCell", quietly = TRUE)}

pbmc4k2 <- post_process(pbmc4k)

SC <- get_supercell(sce_sc, element = "object")

purity <- SuperCell::supercell_purity(colLabels(pbmc4k2), SC$membership, method = "entropy")

head(purity)

summary(purity)
```
