# Coarse-graining of large single-cell data into metacells using SuperCell {#supercell}


Here, we use the `pbmc4k` dataset as presented in [Batch correction](https://yulab-smu.top/sclet/batch-correction.html#preprocess). 


## `RunSuperCell` 


The `RunSuperCell()` function is a wrapper function to run [SuperCell](https://github.com/GfellerLab/SuperCell).


```{r run-SC}
pbmc <- RunSuperCell(pbmc4k, gamma = 5)
pbmc
```


The output of `RunSuperCell()` is a `SingleCellExperiment` object that stores gene expression matrix of metacells. 

We can use it for downstram analysis.

```{r supercell, fig.width=5, fig.height=4}
post_process <- function(pbmc) {
    pbmc <- NormalizeData(pbmc)
    pbmc <- FindVariableFeatures(pbmc)
    pbmc <- ScaleData(pbmc)
    pbmc <- runPCA(pbmc, subset_row = VariableFeatures(pbmc), exprs_values = "scaled")
    
    pbmc <- FindNeighbors(pbmc, dims = 1:10)
    pbmc <- FindClusters(pbmc)
    pbmc <- RunUMAP(pbmc)
    return(pbmc)
}

pbmc2 <- post_process(pbmc)

library(ggsc)
sc_dim(pbmc2, reduction="UMAP") + sc_dim_geom_label()
```


## Estimate SuperCell purity

```{r SC-purity}

pbmc4k2 <- post_process(pbmc4k)

SC <- metadata(pbmc2)$SuperCell

purity <- SuperCell::supercell_purity(pbmc4k2$label, SC$membership, method = 'entropy')

head(purity)

summary(purity)
```
