# SVP
Robust analysis of 'gene set' activity in spatial or single-cell data.

```{r}
library(SpatialExperiment)
library(SingleCellExperiment)
library(scuttle)
library(SVP)
library(ggplot2)
library(ggsc)
library(sclet)
```

## runSGSA
Calculate the activity of gene sets in spatial or single-cell data with restart walk with restart and hyper test weighted.
sceSubPbmc is a small SingleCellExperiment data set from pbmck3 which contains 1304 genes and 800 cells (extract randomly).
```{r}
data(sceSubPbmc)
sceSubPbmc <- NormalizeData(sceSubPbmc)
# the using runMCA to perform MCA (Multiple Correspondence Analysis)
# this is refer to the CelliD, but we using the Eigen to speed up.
# You can view the help information of runMCA using ?runMCA.
sceSubPbmc <- runMCA(sceSubPbmc, assay.type = 'logcounts')
```


CellCycle.Hs is the S and G2M gene list are from the Seurat which refer to this article (doi:10.1126/science.aad050), the G1 gene list is from the G1_PHASE of Human Gene Set in MSigDB, but remove the duplicated records with S and G2M gene list.
```{r}
# Next, we can calculate the activity score of gene sets provided.
# Here, we use the Cell Cycle gene set from the Seurat 
# You can use other gene set, such as KEGG pathway, GO, Hallmark of MSigDB
# or TFs gene sets etc.
data(CellCycle.Hs)
sceSubPbmc <- runSGSA(sceSubPbmc, gset.idx.list = CellCycle.Hs, gsvaExp.name = 'CellCycle')
# Then a SVPE class which inherits SingleCellExperiment, is return.
sceSubPbmc
```

```{r}
# You can obtaion the score matrix by following the commond
sceSubPbmc |> gsvaExp('CellCycle') 
sceSubPbmc |> gsvaExp("CellCycle") |> assay() |> t() |> head()
```

```{r}
# Then you can use the ggsc or other package to visulize
# and you can try to use the findMarkers of scran or other packages to identify
# the different gene sets.   
sceSubPbmc <- sceSubPbmc |> 
                scater::runPCA(assay.type = 'logcounts', ntop = 600) |>
                scater::runUMAP(dimred = 'PCA')
# withReducedDim = TRUE, the original reducetion results from original gene features
# will be add the colData in the sce.cellcycle.
sce.cellcycle <- sceSubPbmc |> gsvaExp('CellCycle', withReducedDim=TRUE)
sce.cellcycle
```

```{r svp-cc-violin, fig.width=10, fig.height=10}
sce.cellcycle |> sc_violin(
                      features = rownames(sce.cellcycle), 
                      mapping = aes(x=seurat_annotations, fill = seurat_annotations)
                   ) + 
                   scale_x_discrete(guide=guide_axis(angle=-45))
```

```{r svp-cc-s-umap, fig.width=10, fig.height=10}
sce.cellcycle |> sc_feature(features= "S", reduction='UMAP')
```

```{r}
library(scran)
cellcycle.test.res <- sce.cellcycle |> findMarkers(
                group = sce.cellcycle$seurat_annotations, 
                test.type = 'wilcox', 
                assay.type = 'affi.score', 
                add.summary = TRUE
                )
cellcycle.test.res$B
```

## runLISA
hpda_spe_cell_dec is the result of runSGSA with HPDA A sample from (doi:10.1038/s41587-019-0392-8). 
Each item in the matrix is the marker gene set activity in the spot. The marker gene set represents the cell type.
We can use svp the investigate if a cell type cluters and if two cell types appear in the same region.
Usage of local indicators of spatial association (LISA) to identify the hotspot in the spatial space
```{r}
data(hpda_spe_cell_dec)

# r$> assay(hpda_spe_cell_dec)[1:5,1:5]
# 5 x 5 sparse Matrix of class "dgCMatrix"
#                                             Spot1        Spot2        Spot3        Spot4        Spot5
# Acinar cells                         3.283881e-04 6.931743e-06 4.820951e-05 1.882577e-03 9.269843e-04
# Cancer clone A                       2.880388e-04 2.013973e-04 9.167013e-05 9.642225e-05 6.704812e-05
# Cancer clone B                       1.569730e-04 2.744810e-04 2.151324e-04 1.502651e-05 4.283322e-06
# Ductal APOL1 high-hypoxic            3.132481e-05 3.902286e-04 4.240874e-05 3.140970e-06 1.464118e-04
# Ductal CRISP3 high-centroacinar like 9.546611e-05 5.025320e-03 2.112512e-03 1.140155e-03 3.873088e-03

svres <- runDetectSVG(hpda_spe_cell_dec, assay.type = 'affi.score', 
                      method = 'moransi', action = 'only') 

# r$> svres |> dplyr::arrange(rank) |> head()
#                                            obs expect.moransi sd.moransi Z.moransi        pvalue          padj rank
# Cancer clone A                       0.7086916    -0.00234192 0.02803920  25.35855 3.616015e-142 7.232030e-141    1
# Acinar cells                         0.6958980    -0.00234192 0.02788076  25.04379 1.020141e-138 1.020141e-137    2
# Cancer clone B                       0.6586104    -0.00234192 0.02803039  23.57985 3.102650e-123 2.068434e-122    3
# mDCs A                               0.5824019    -0.00234192 0.02760394  21.18334 6.801396e-100  3.400698e-99    4
# Ductal CRISP3 high-centroacinar like 0.5605955    -0.00234192 0.02794187  20.14673  1.437344e-90  5.749375e-90    5
# Endothelial cells                    0.5362475    -0.00234192 0.02757573  19.53128  2.976221e-85  9.920735e-85    6
```


```{r}
lisa.res12 <- hpda_spe_cell_dec |>
   runLISA(
     features = c(1, 2, 3), 
     assay.type = 'affi.score',
     weight.method = "knn",
     k = 10,
     action = 'get',
   )
lisa.res12
lisa.res12[['Acinar cells']] |> head()
lisa.res12[["Cancer clone A"]] |> head()
colData(hpda_spe_cell_dec)$`cluster.test.Cancer.A` <- lisa.res12[["Cancer clone A"]] |>
dplyr::pull(cluster.test)
colData(hpda_spe_cell_dec)$`cluster.test.Acinar` <- lisa.res12[["Acinar cells"]] |>
dplyr::pull(cluster.test)
colData(hpda_spe_cell_dec)$`cluster.test.Cancer.B` <- lisa.res12[["Cancer clone B"]] |>
dplyr::pull(cluster.test)
```

```{r svp-hpda-spatial-1, fig.width=10, fig.height=10}
p1 <- sc_spatial(hpda_spe_cell_dec,
            features = rownames(hpda_spe_cell_dec),
            mapping = aes(x=x, y=y, color=cluster.test.Cancer.A),
            plot.pie = T,
            pie.radius.scale = .8,
            bg_circle_radius = 1.1,
            color=NA,
            linewidth=2
) + scale_color_manual(values=c("black", "white"))
p1
```


```{r svp-hpda-spatial-2, fig.width=10, fig.height=10}
f1 <- sc_spatial(hpda_spe_cell_dec, features="Cancer clone A",
            mapping=aes(x = x, y = y),
            pointsize=10) + 
        geom_scattermore2(mapping = aes(bg_color=cluster.test.Cancer.A, 
                          subset=cluster.test.Cancer.A=="High"),
                          bg_line_width = .16,
                          gap_line_width = .14,
                          pointsize = 7) +
        scale_bg_color_manual(values=c('black'))
f1
```


```{r svp-hpda-spatial-3, fig.width=10, fig.height=10}

f2 <- sc_spatial(hpda_spe_cell_dec, features="Acinar cells",
            mapping=aes(x=x,y=y),
            pointsize=10) +
        geom_scattermore2(mapping = aes(bg_color=cluster.test.Acinar, subset=cluster.test.Acinar=="High"),
                          bg_line_width = .16,
                          gap_line_width = .14,
                          pointsize = 7) +
        scale_bg_color_manual(values=c('black'))
f2
```



```{r svp-hpda-spatial-4, fig.width=10, fig.height=10}
f3 <- sc_spatial(hpda_spe_cell_dec, features="Cancer clone B",
            mapping=aes(x=x,y=y),
            pointsize=10) +
        geom_scattermore2(mapping = aes(bg_color=cluster.test.Cancer.B, subset=cluster.test.Cancer.B=="High"),
                          bg_line_width = .18,
                          gap_line_width = .14,
                          pointsize = 8) +
        scale_bg_color_manual(values=c('black'))
f3
```


## runLOCALBV
explore the local bivariate relationship in the spatial space.
```{r eval = FALSE}
res1 <- hpda_spe_cell_dec |> runLOCALBV(
          features1 = 'Cancer clone A', 
          features2 = 'Cancer clone B', 
          assay.type='affi.score'
        )
res1
res1[['Cancer clone A_VS_Cancer clone B']] |> head()


#        LocalLee           Gi       E.Gi       Var.Gi       Z.Gi Pr (z > E(Gi)) cluster.no.test cluster.test
# Spot1 0.2656572 0.0007993283 0.00234192 1.685012e-06 -1.1883644      0.8826551             Low       NoSign
# Spot2 0.1928726 0.0006520670 0.00234192 3.060135e-06 -0.9660035      0.8329788             Low       NoSign
# Spot3 0.2098906 0.0007134499 0.00234192 2.544445e-06 -1.0209003      0.8463492             Low       NoSign
# Spot4 0.2609793 0.0007131499 0.00234192 2.545557e-06 -1.0208654      0.8463409             Low       NoSign
# Spot5 0.1769749 0.0007300670 0.00234192 3.059710e-06 -0.9214789      0.8215998             Low       NoSign
# Spot6 0.1736672 0.0004656579 0.00234192 2.175086e-06 -1.2721999      0.8983489             Low       NoSign
```
