# Trajectory inference and RNA velocity {#trajectory}

According to [Wikipedia](https://en.wikipedia.org/wiki/Trajectory_inference):

> Trajectory inference or pseudotemporal ordering is a computational technique used in single-cell transcriptomics to determine the pattern of a dynamic process experienced by cells and then arrange cells based on their progression through the process.

It should be noted that all trajectory inference methods are based on the current transcriptional state and the distance symmetry between cells. That is to say, the distance from cell A to cell B is equal to the distance from cell B to cell A. Therefore, the direction of the inferred trajectory is either missing (requiring user specification) or unreliable (using default settings).


## Slingshot: cell lineage and pseudotime inference {#slingshot}


```{r}
pancreas <- readRDS("data/pancreas_sub_sce.rds")

panc <- RunSlingshot(sce = pancreas, group = "SubCellType", reduction = "UMAP")
```



### Lineage plot
```{r lineage_plot, fig.width=7, fig.height=8}
p1 <- lineage_plot(panc, group = "SubCellType") + 
  sc_dim_geom_label(
    geom = shadowtext::geom_shadowtext,
    color='black', 
    bg.color='white'
  )


## displayed selected lineages
p2 <- lineage_plot(panc, group = "SubCellType", lineages = c("Lineage1", "Lineage2"))  + 
  sc_dim_geom_label(
    geom = shadowtext::geom_shadowtext,
    color='black', 
    bg.color='white'
  )

plot_list(`All Lineages` = p1, `Selected Lineages` = p2, ncol=1)
```

### Pseudotime plot

```{r pseudotimeplot, fig.width=12, fig.height=10}
pp1 <- pseudo_plot(panc, reduction = "UMAP", lineage = "Lineage1")
pp2 <- pseudo_plot(panc, reduction = "UMAP", lineage = "Lineage2")
pp_all <- pseudo_plot(panc, reduction = "UMAP")

plot_list(pp1, pp2, pp_all, design="AABB\nCCCC")
```


### Expression trends in different cell trajectories

```{r genecurve, fig.width=14, fig.height=8}
curve_genes <- c("Mrpl15", "Tram1", "Ncoa2")

## 默认选取smooth方法直接拟合得到的基因表达值结果来画图
p_curve <- genecurve_plot(panc, features = curve_genes, method = "smooth")

## 也可以用GAM计算预测的基因表达值来画图
p_curve1 <- genecurve_plot(panc, features = curve_genes, method = "gam")

plot_list(p_curve, p_curve1, ncol=1)
```



### Heatmap

```{r pseudoheatmap, fig.width=10, fig.height=16}
heatmap_genes <- head(unique(rownames(panc)), 10)

## 默认按照拟时间排序
hmp1 <- pseudo_heatmap(panc, features = heatmap_genes, lineage = "Lineage2")

## 可以手动对基因或细胞设置聚类(cluster = TRUE)，此时会打乱pseudotime的排序
hmp2 <- pseudo_heatmap(panc, features = heatmap_genes, lineage = "Lineage2", cluster_columns = TRUE)

## 对基因聚类
hmp3 <- pseudo_heatmap(panc, features = heatmap_genes, lineage = "Lineage2", cluster_rows = TRUE, sort = FALSE)

## 默认排序
hmp4 <- pseudo_heatmap(panc, features = heatmap_genes, lineage = "Lineage2", cluster_rows = FALSE, sort = TRUE)

## 组合起来看就能看到两个参数的设定对行顺序的影响
aplot::plot_list(hmp1, hmp2, hmp3, hmp4, ncol=1)
```





## RNA velocity {#Velocity}

RNA metabolic process:

+ RNA is synthesized at a rate of alpha 
+ RNA is spliced to remove introns and form mature mRNA at rate of beta
+ Mature mRNA is degraded after functioning at rate of gamma

The disruption of the equilibrium between unspliced and spliced RNA indicates whether a gene is in an induced or repressed state. 

With RNA-seq, especially the Poly-A enrichment method, the propotion of unspliced RNA obtained is relatively small. Currently, there are also some techinques capable of enriching nascent RNA in single cells, such as [scSLAM-seq](https://www.nature.com/articles/s41586-019-1369-y) and s[cNT-seq](https://www.nature.com/articles/s41592-020-0935-4).


Although some researchers are skeptical about RNA velocity, it has received significant attention because it can automatically detect the direction of trajectories.


[scVelo](https://scvelo.readthedocs.io/) supports a full dynamical model and various of utility functions [@bergen_generalizing_2020]. It only supports Python and can be run in R using `r Biocpkg("velociraptor")`. 

### Example data

> This example is adopted from <https://www.bioconductor.org/packages/devel/bioc/vignettes/velociraptor/inst/doc/velociraptor.html>.


```{r velocity-data}
library(scRNAseq)
sce <- HermannSpermatogenesisData()
sce

# first 500 cells for demonstration
sce <- sce[, 1:500]
```

```{r velocity-hvgs}
library(scuttle)
sce <- logNormCounts(sce, assay.type=1)

library(sclet)
sce <- FindVariableFeatures(sce, method="scran")
hvgs <- VariableFeatures(sce, method="scran")
```

### Run velocity analysis

```{r velocity}
library(velociraptor)

velo.out <- scvelo(
  sce, subset.row=hvgs, assay.X="spliced",
  scvelo.params=list(neighbors=list(n_neighbors=30L))
)

velo.out
```

### Visualize velocity-based trajectory

```{r velocity-plot, fig.width=12, fig.height=8.5}
library(scater)

set.seed(100)
sce <- runPCA(sce, subset_row=hvgs)
sce <- runTSNE(sce, dimred="PCA", perplexity = 30)

sce$velocity_pseudotime <- velo.out$velocity_pseudotime

embedded <- embedVelocity(reducedDim(sce, "TSNE"), velo.out)
grid.df <- gridVectors(sce, embedded, use.dimred = "TSNE")


require(ggsc)
sc_dim(sce, reduction='TSNE', mapping=aes(color=velocity_pseudotime)) + 
    scale_color_viridis_c() +
    geom_segment(data=grid.df, 
        mapping=aes(x=start.1, y=start.2, 
            xend=end.1, yend=end.2, colour=NULL), 
        arrow=arrow(length=unit(0.05, "inches")), alpha=.6)
```
