# Trajectory Inference and RNA Velocity {#trajectory}

Static clustering tells us which cell states are present, but many biological systems are dynamic. Development, activation, differentiation, and response to perturbation all involve cells moving through transitional states rather than staying in discrete categories.

This chapter therefore shifts from classification to dynamics. We begin with trajectory inference based on transcriptional similarity, then move to RNA velocity and related methods that add directional information to the analysis. Together, these approaches help answer not only *what* a cell is, but also *where it may be going next*.


## Slingshot: cell lineage and pseudotime inference {#slingshot}


```{r}
panc <- readRDS("data/pancreas_sub_sce.rds")

panc <- RunSlingshot(sce = panc, group = "SubCellType", reduction = "UMAP")
```

`RunSlingshot()` writes its result back into the unified state layer, so downstream helpers such as `plot_lineage()` and `plot_pseudotime()` can consume the recorded trajectory object directly. If needed, you can also check `has_trajectory(panc)` or inspect `get_trajectory(panc)`.

If you keep multiple trajectory runs in the same object, give them different `name` values and inspect them explicitly with `get_trajectory(panc, id = "your_id")`.




### Lineage plot
```{r lineage_plot, fig.width=7, fig.height=8}
p1 <- plot_lineage(panc, group = "SubCellType") + 
  sc_dim_geom_label(
    geom = shadowtext::geom_shadowtext,
    color='black', 
    bg.color='white'
  )


## displayed selected lineages
p2 <- plot_lineage(panc, group = "SubCellType", lineages = c("Lineage1", "Lineage2"))  + 
  sc_dim_geom_label(
    geom = shadowtext::geom_shadowtext,
    color='black', 
    bg.color='white'
  )

plot_list(`All Lineages` = p1, `Selected Lineages` = p2, ncol=1)
```

### Pseudotime plot

```{r pseudotimeplot, fig.width=12, fig.height=10}
pp1 <- plot_pseudotime(panc, reduction = "UMAP", lineage = "Lineage1")
pp2 <- plot_pseudotime(panc, reduction = "UMAP", lineage = "Lineage2")
pp_all <- plot_pseudotime(panc, reduction = "UMAP")

plot_list(pp1, pp2, pp_all, design="AABB\nCCCC")
```


### Expression trends in different cell trajectories

```{r genecurve, fig.width=14, fig.height=8}
curve_genes <- c("Mrpl15", "Tram1", "Ncoa2")

## 默认选取smooth方法直接拟合得到的基因表达值结果来画图
p_curve <- plot_genecurve(panc, features = curve_genes, method = "smooth")

## 也可以用GAM计算预测的基因表达值来画图
p_curve1 <- plot_genecurve(panc, features = curve_genes, method = "gam")

plot_list(p_curve, p_curve1, ncol=1)
```



### Heatmap

```{r pseudoheatmap, fig.width=10, fig.height=16}
heatmap_genes <- head(unique(rownames(panc)), 10)

## 默认按照拟时间排序
hmp1 <- plot_pseudo_heatmap(panc, features = heatmap_genes, lineage = "Lineage2")

## 可以手动对基因或细胞设置聚类(cluster = TRUE)，此时会打乱pseudotime的排序
hmp2 <- plot_pseudo_heatmap(panc, features = heatmap_genes, lineage = "Lineage2", cluster_columns = TRUE)

## 对基因聚类
hmp3 <- plot_pseudo_heatmap(panc, features = heatmap_genes, lineage = "Lineage2", cluster_rows = TRUE, sort = FALSE)

## 默认排序
hmp4 <- plot_pseudo_heatmap(panc, features = heatmap_genes, lineage = "Lineage2", cluster_rows = FALSE, sort = TRUE)

## 组合起来看就能看到两个参数的设定对行顺序的影响
plot_list(hmp1, hmp2, hmp3, hmp4, ncol=1)
```



## Diffusion Map for complex trajectories {#diffusion-map}

For very continuous developmental systems, PCA and UMAP are often convenient but not always topology-faithful. `RunDiffusionMap()` adds a diffusion-based representation that can be tracked like any other reduction in `sclet`, which is useful when you want a more trajectory-aware geometry before moving on to downstream lineage tools.

```{r diffusion-map, eval = eval_chunks && requireNamespace("destiny", quietly = TRUE)}
panc_dm <- panc

if (!"logcounts" %in% SummarizedExperiment::assayNames(panc_dm) &&
    "counts" %in% SummarizedExperiment::assayNames(panc_dm)) {
  panc_dm <- NormalizeData(panc_dm)
}

if (!"PCA" %in% SingleCellExperiment::reducedDimNames(panc_dm)) {
  dm_features <- rownames(panc_dm)[seq_len(min(500, nrow(panc_dm)))]
  dm_layer <- if ("logcounts" %in% SummarizedExperiment::assayNames(panc_dm)) "logcounts" else "counts"
  panc_dm <- RunPCA(panc_dm, subset_row = dm_features, layer = dm_layer)
}

panc_dm <- RunDiffusionMap(panc_dm, n_components = 10, reduction = "PCA")
head(SingleCellExperiment::reducedDim(panc_dm, "DM")[, 1:3])
```

```{r diffusion-map-plot, fig.width=8, fig.height=6, eval = eval_chunks && requireNamespace("destiny", quietly = TRUE)}
DimPlot(panc_dm, reduction = "DM", mapping = ggplot2::aes(color = SubCellType))
```






The velocity and CellRank examples in this section are executed during book rendering when the required Python environment can be provisioned through `basilisk`. This means the documentation now also exercises the cross-language trajectory stack in CI instead of only describing it conceptually.

RNA metabolic process:

+ RNA is synthesized at a rate of alpha 
+ RNA is spliced to remove introns and form mature mRNA at rate of beta
+ Mature mRNA is degraded after functioning at rate of gamma

The disruption of the equilibrium between unspliced and spliced RNA indicates whether a gene is in an induced or repressed state. 

With RNA-seq, especially the Poly-A enrichment method, the propotion of unspliced RNA obtained is relatively small. Currently, there are also some techinques capable of enriching nascent RNA in single cells, such as [scSLAM-seq](https://www.nature.com/articles/s41586-019-1369-y) and s[cNT-seq](https://www.nature.com/articles/s41592-020-0935-4).


Although some researchers are skeptical about RNA velocity, it has received significant attention because it can automatically detect the direction of trajectories.


[scVelo](https://scvelo.readthedocs.io/) supports a full dynamical model and various of utility functions [@bergen_generalizing_2020]. It only supports Python and can be run in R using `r Biocpkg("velociraptor")`. 

### Example data

> This example is adopted from <https://www.bioconductor.org/packages/devel/bioc/vignettes/velociraptor/inst/doc/velociraptor.html>.


```{r velocity-data}
library(scRNAseq)
sce <- HermannSpermatogenesisData()
sce

# first 500 cells for demonstration
sce <- sce[, 1:500]
```

```{r velocity-hvgs}
library(sclet)
sce <- NormalizeData(sce, assay = "spliced")
sce <- FindVariableFeatures(sce, method="scran")
hvgs <- VariableFeatures(sce, method="scran")
sce <- RunPCA(sce, subset_row = hvgs, layer = "logcounts")
sce <- FindNeighbors(sce, dims = 1:20)
sce <- FindClusters(sce)
sce$cluster <- as.character(SingleCellExperiment::colLabels(sce))
sce <- RunUMAP(sce, dims = 1:20)
```

### Run velocity analysis

`sclet` wraps the RNA velocity inference via `RunVelocity()`. Under the hood, it delegates the heavy lifting to `scVelo` (managed safely via the `basilisk`/`velociraptor` Python sandbox) and smoothly registers the outputs to the `velocity` analysis state. This eliminates the need for manual environment switching.

```{r velocity, eval=eval_chunks && requireNamespace("velociraptor", quietly = TRUE)}
# Run Velocity (Requires 'spliced' and 'unspliced' assays to be present)
sce <- RunVelocity(sce, subset_row = hvgs, assay.X = "spliced", use.dimred = "PCA")

# Check the provenance state to confirm successful execution
get_velocity(sce)
```

### Visualize velocity-based trajectory

Once the `velocity` state is recorded, visualizing it is a breeze. You can use `VelocityPlot()` to draw the dynamic stream or arrows over any active dimensionality reduction space (like UMAP or t-SNE).

```{r velocity-plot, eval=eval_chunks && requireNamespace("velociraptor", quietly = TRUE)}
# Project the velocity stream onto the UMAP embedding
VelocityPlot(sce, reduction = "UMAP")
```

## CellRank: Advanced Fate Mapping {#cellrank}

When standard trajectory tools (like Slingshot) struggle with complex, cyclic, or highly heterogeneous developmental processes, **CellRank** shines. CellRank elegantly combines directed RNA velocity data with undirected transcriptomic similarity to robustly infer terminal states and cellular fate probabilities.

`sclet` brings the full power of `CellRank` directly to your SCE object, again utilizing the isolated `basilisk` Python environment to prevent dependency conflicts.

### Run CellRank

`RunCellFate()` is the user-facing entry point for fate mapping in `sclet`. At the moment it dispatches to `CellRank`, automatically senses the upstream velocity run, and records the resulting terminal states, named fate probabilities, and lineage drivers in the shared state contract.

```{r cellrank-run, eval=eval_chunks && requireNamespace("velociraptor", quietly = TRUE) && requireNamespace("basilisk", quietly = TRUE) && requireNamespace("reticulate", quietly = TRUE)}
# Compute transition matrix, terminal states, and absorption probabilities
sce <- RunCellFate(sce, reduction = "PCA", cluster_key = "cluster")

# Inspect the CellRank analysis state
get_cellrank(sce)
```

The resulting terminal state assignments and directional fate probabilities are written directly into `colData(sce)`. In addition, `sclet` now provides fate-specific plotting helpers following the `plot_<module>_<type>` convention, so you do not need to manually reconstruct the plotting data frame after every run.

```{r cellrank-plot, fig.width=12, fig.height=5.5, eval=eval_chunks && requireNamespace("velociraptor", quietly = TRUE) && requireNamespace("basilisk", quietly = TRUE) && requireNamespace("reticulate", quietly = TRUE)}
p_prob <- plot_fate_probability(sce, fate = 1, reduction = "UMAP")
p_term <- plot_fate_terminal_states(sce, reduction = "UMAP")

plot_list(`Fate Probability` = p_prob, `Terminal States` = p_term)
```

```{r cellrank-driver-trends, fig.width=11, fig.height=7, eval=eval_chunks && requireNamespace("velociraptor", quietly = TRUE) && requireNamespace("basilisk", quietly = TRUE) && requireNamespace("reticulate", quietly = TRUE)}
cr <- get_cellrank(sce)
driver_lineage <- if (!is.null(cr$artifacts$fate_probability_names)) {
  cr$artifacts$fate_probability_names[[1]]
} else {
  cr$fate_probability_names[[1]]
}

plot_fate_driver_trends(sce, lineage = driver_lineage, top_n = 4)
```
