# Enhanced Visualization

Once a dataset has been processed, plotting becomes the main way we inspect whether the workflow is behaving as expected. We need to check clusters, marker expression, sample composition, and structured summaries across groups, often with many small figure adjustments along the way.

`sclet` provides an enhanced visualization ecosystem tightly integrated with the `ggsc` package. It offers high-level plotting functions that operate directly on `SingleCellExperiment` objects while still returning standard `ggplot2` objects, so you keep both convenience and full downstream customizability.

## Loading data

```{r vis-load, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r vis-data}
library(sclet)
f <- "data/pbmc-3k-sce.qs"
pbmc <- qs::qread(f)
```

## Dimensionality Reduction Plots

`sclet` wraps `ggsc::sc_dim` and `ggsc::sc_feature` to provide `CellDimPlot` and `FeatureDimPlot`. These functions automatically resolve the active reduction and layer from the object's state.

### CellDimPlot

`CellDimPlot` is used to visualize categorical cell metadata on a reduced dimensional space (e.g., UMAP, PCA).

```{r cell-dim-plot, fig.width=12, fig.height=5}
library(ggplot2)
library(aplot)

# Visualize clustering results
p1 <- CellDimPlot(pbmc, group.by = "ident") + ggtitle("Clusters")

# Visualize batch or other metadata
# Assuming we have a 'label' column in colData
if ("label" %in% colnames(colData(pbmc))) {
    p2 <- CellDimPlot(pbmc, group.by = "label") + ggtitle("Cell Types")
    plot_list(p1, p2)
} else {
    p1
}
```

### FeatureDimPlot

`FeatureDimPlot` is used to visualize continuous variables, such as gene expression levels, on the reduced dimensional space.

```{r feature-dim-plot, fig.width=12, fig.height=5}
# Visualize expression of specific marker genes
FeatureDimPlot(pbmc, features = c("MS4A1", "CD8A"))
```

## Expression Heatmaps

`GroupHeatmap` provides a convenient way to visualize the average expression of features across different cell groups using `scater::plotGroupedHeatmap`.

```{r group-heatmap, fig.width=8, fig.height=6}
# Define a set of marker genes
markers <- c("IL7R", "CD14", "LYZ", "MS4A1", "CD8A", "FCGR3A", "MS4A7")

# Plot average expression by cluster
GroupHeatmap(pbmc, features = markers, group.by = "ident", scale = TRUE)
```

## Cell Statistics

`CellStatPlot` is useful for visualizing the composition of cell groups (e.g., cell type proportions across different conditions).

```{r cell-stat-plot, fig.width=8, fig.height=5}
# If we have multiple samples or batches, we can plot the proportion of each cluster
# Here we just use a simulated 'batch' column for demonstration if one doesn't exist
if (!"batch" %in% colnames(colData(pbmc))) {
    set.seed(123)
    pbmc$batch <- sample(c("Sample1", "Sample2"), ncol(pbmc), replace = TRUE)
}

CellStatPlot(pbmc, stat.by = "ident", group.by = "batch", position = "fill") +
    ylab("Proportion") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

These visualization functions are designed to minimize boilerplate code while remaining fully compatible with `ggplot2` grammar (`+ theme()`, `+ scale_color_manual()`, etc.).
