# Enhanced Visualization

Once a dataset has been processed, plotting becomes the main way we inspect whether the workflow is behaving as expected. We need to check clusters, marker expression, sample composition, and structured summaries across groups, often with many small figure adjustments along the way.

`sclet` provides an enhanced visualization ecosystem tightly integrated with the `ggsc` package. It offers high-level plotting functions that operate directly on `SingleCellExperiment` objects while still returning standard `ggplot2` objects, so you keep both convenience and full downstream customizability.

## Loading data

```{r vis-load, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r vis-data}
library(sclet)
f <- "data/pbmc-3k-sce.rds"
pbmc <- readRDS(f)
```

## Dimensionality Reduction Plots

`sclet` wraps `ggsc::sc_dim` and `ggsc::sc_feature` to provide `CellDimPlot` and `FeatureDimPlot`. These functions automatically resolve the active reduction and layer from the object's state.

### CellDimPlot

`CellDimPlot` is used to visualize categorical cell metadata on a reduced dimensional space (e.g., UMAP, PCA).

```{r cell-dim-plot, fig.width=12, fig.height=5}
library(ggplot2)
library(aplot)

# Visualize clustering results
p1 <- CellDimPlot(pbmc, group.by = "ident") + ggtitle("Clusters")

# Visualize batch or other metadata
# Assuming we have a 'label' column in colData
if ("label" %in% colnames(colData(pbmc))) {
    p2 <- CellDimPlot(pbmc, group.by = "label") + ggtitle("Cell Types")
    plot_list(p1, p2)
} else {
    p1
}
```

### FeatureDimPlot

`FeatureDimPlot` is used to visualize continuous variables, such as gene expression levels, on the reduced dimensional space.

```{r feature-dim-plot, fig.width=12, fig.height=5}
# Visualize expression of specific marker genes
FeatureDimPlot(pbmc, features = c("MS4A1", "CD8A"))
```

## Expression Heatmaps

`GroupHeatmap` provides a convenient way to visualize the average expression of features across different cell groups using `scater::plotGroupedHeatmap`.

```{r group-heatmap, fig.width=8, fig.height=6}
# Define a set of marker genes
markers <- c("IL7R", "CD14", "LYZ", "MS4A1", "CD8A", "FCGR3A", "MS4A7")

# Plot average expression by cluster
GroupHeatmap(pbmc, features = markers, group.by = "ident", scale = TRUE)
```

## Cell Statistics

`CellStatPlot` is useful for visualizing the composition of cell groups (e.g., cell type proportions across different conditions).

```{r cell-stat-plot, fig.width=8, fig.height=5}
# If we have multiple samples or batches, we can plot the proportion of each cluster
# Here we just use a simulated 'batch' column for demonstration if one doesn't exist
if (!"batch" %in% colnames(colData(pbmc))) {
    set.seed(123)
    pbmc$batch <- sample(c("Sample1", "Sample2"), ncol(pbmc), replace = TRUE)
}

CellStatPlot(pbmc, stat.by = "ident", group.by = "batch", position = "fill") +
    ylab("Proportion") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

These visualization functions are designed to minimize boilerplate code while remaining fully compatible with `ggplot2` grammar (`+ theme()`, `+ scale_color_manual()`, etc.).

## Module-Specific Visualization

Beyond the general-purpose plotting functions, `sclet` provides specialized visualization functions for each analysis module. These consume the typed state records directly, so you never need to manually extract matrices or metadata columns.

**Note:** The examples below show the intended API surface. They use `eval=FALSE` because the PBMC demo object used in this chapter has not been processed through trajectory, velocity, spatial, or CCI analysis pipelines. The functions themselves are fully implemented and ready to use once the corresponding module has been run.

### Trajectory and Fate

```{r viz-fate, eval=FALSE}    # requires RunVelocity() / RunCellRank() output
# Trajectory overview panel
plot_trajectory_overview(sce, panels = c("velocity", "terminal_states", "fate_probability"))

# Fate probability on embedding
plot_fate_probability(sce, fate = 1, reduction = "UMAP")

# Terminal state assignments
plot_fate_terminal_states(sce, reduction = "UMAP")

# Lineage driver trends
plot_fate_driver_trends(sce, lineage = "Lineage1", top_n = 4)

# Velocity latent time
plot_velocity_latent_time(sce, reduction = "UMAP")
```

### Program and Regulon Activity

Gene set scores and regulons share a unified visualization layer. The `source` parameter lets you switch between "geneset_scoring" and "scenic" without changing the plot call.

```{r viz-program, eval=FALSE}
# Single program on embedding
plot_program_activity(sce, program = "T_cell", source = "geneset_scoring")

# Multi-program heatmap across groups
plot_program_heatmap(sce, programs = c("T_cell", "B_cell", "Myeloid"))

# Program dotplot (size = percent expressed, color = mean activity)
plot_program_dotplot(sce, programs = c("T_cell", "B_cell", "Myeloid"), group.by = "colLabels")

# SCENIC regulon on embedding
plot_scenic_activity(sce, regulon = "STAT1(+)")
```

### Perturbation Priority

```{r viz-perturbation, eval=FALSE}
# Ranked bar chart of perturbation AUC per cell type
plot_perturbation_ranking(sce)
```

### Spatial Analysis

```{r viz-spatial, eval=FALSE}
# Deconvolution composition bar chart
plot_spatial_deconvolution(sce)

# Cell type × spot proportion heatmap
plot_spatial_composition(sce, n = 10)
```

### Cell-Cell Communication

```{r viz-cci, eval=FALSE}
plot_cci_bubble(sce, type = "interaction")
plot_cci_sigmoid(sce)
plot_cci_network(sce)
```

### Reference Mapping

```{r viz-mapping, eval=FALSE}
# Label transfer summary heatmap
plot_reference_label_transfer_heatmap(sce, group.by = "colLabels")

# Per-label confidence distribution
plot_reference_label_confidence(sce)
```
